# Lavaan Lab 1: Path Analysis Model

```{r set-options, echo=FALSE, cache=FALSE, message=FALSE}
library(knitr)
options(width = 1800)
opts_chunk$set(cache=TRUE)
```

In this lab, we will learn how to: 

+ install a package called lavaan in R
+ perform path analysis using the lavaan package

## Reading-In and Working With Realistic Datasets In R

#### To begin, we will read the file that we will use for our SEM lab (eatingDisorderSimData.csv).

Try running this function, as written:

```{r, eval = FALSE}
file.choose()
```

Using the GUI (graphical user interface) window that pops up,
select the file eatingDisorderSimData.csv

This should produce a file path like this (note: below is a Mac version):

```{r, eval = FALSE}
/Your/File/Path/eatingDisorderSimData.csv
```

You can copy this path into the read.csv and put it in the file = argument of the function: 

+ read.csv() is a function for reading in .csv files.
+ Assign the name labData to the dataset in R using <-

```{r, eval = FALSE}
labData <- read.csv(file = "/Users/gejiang/Box Sync/MacSync/Teaching/590SEM/Spring 2022/Week 4/R/eatingDisorderSimData.csv", header = TRUE)
```

Important Argument:
header =

+ if header = TRUE, indicates that your dataset has column names that are to be read in separate from the data.
+ if header = FALSE, indicates that your dataset does NOT have column names, and therefore the first row of the dataset should be read as data.

#### Or you could NEST the file.choose() function inside the read.csv function

```{r, eval = FALSE}
labData <- read.csv(file = file.choose(), header = T)
```

Because file.choose() returns the file path, putting this inside the read.csv function is the same as writing the path inside the function!

#### Pros and Cons of writing the full file path vs. using read.csv(file = file.choose(), header = T)

If you write down the full file path and put it in the function, then the next time you run this R script you can easily read in your data without searching through your directories and folders.

However, if you move your file to a different folder in the future, you'll need to change the directory path in your R script.

file.choose() is very easy and user-friendly.

Using this method allows you to find your datafile even if you've moved it to a different folder.

However, it is slightly more effortful to go in and select your folder each time.

#### Gabriella recommends:

Set your working directory to the directory that contains the dataset, and simply load your data by typing the name of the .csv file: 

```{r, eval = T}
setwd("~/Box Sync/MacSync/Teaching/590SEM/Spring 2022/Week 4/R")

labData <- read.csv(file = "eatingDisorderSimData.csv", header = T, sep = ",")
```

This serves to save all your future analyses in your working directory. 

read.csv() is related to a broader function called read.table.

The read.table function has a sep = argument
sep = 

+ If sep = "," this indicates a comma-separated (.csv) file
+ If sep = " " this indicates a tab-delimited ("white space" delimited) file, such as a .txt

#### Finally, point and click always works...

```{r, eval = FALSE}
library(readr)
eatingDisorderSimData <- read_csv("eatingDisorderSimData.csv")
View(eatingDisorderSimData)
```


<!------------------------------>
## Sample Covariance Matrices using the cov() function 
<!------------------------------>

#### Quick review:

```{r, eval = T}
str(labData) #structure
head(labData) #first few lines
colnames(labData) #column names
```

+ How many observations are in this dataset?
+ Number of observations = number of rows, with 1 person per row 

```{r, eval = T}
nrow(labData) #1339
```

let's save this number as n
```{r, eval = T}
n <- nrow(labData)
```

Let's look at the sample covariance matrix of these variables using the cov() function:

```{r, eval = T}
cov(labData)
```

let's save this sample cov as capital S:
```{r, eval = T}
S = cov(labData)
```

If we wanted, we could look at a subset of the dataset, e.g.,:

```{r, eval = T}
cov(labData[,c("BMI", "SelfEsteem", "Accu")])
```

This is often useful if our analysis will only contain certain variables.

If only two variables:

```{r, eval = T}
cov(labData$BMI, labData$SelfEsteem)
```

If only one variable (variance):

```{r, eval = T}
cov(labData$BMI, labData$BMI)
```

<!------------------------------>
## Installing Packages 
<!------------------------------>

We will mostly be using the lavaan package to perform SEM analyses, so let's use the install.packages() function to install it first

```{r, eval = F}
install.packages("lavaan")
```

lavaan stands for LAtent VAriable ANalysis using R.

lavaan website: http://lavaan.ugent.be

Check out the tutorials and examples!

<!------------------------------>
## Loading Packages (Libraries) That You Have Installed 
<!------------------------------>

AFTER YOU'VE INSTALLED A PACKAGE ONE TIME, YOU DON'T HAVE TO EVER INSTALL IT AGAIN, UNLESS YOU DELETE AND REINSTALL R FOR SOME REASON.

HOWEVER, NOW THAT THESE FUNCTIONS ARE INSTALLED IN R ON YOUR MACHINE, YOU MUST LOAD THE LIBRARY EVERY TIME YOU OPEN R AND WISH TO USE IT.

To do this, use the library() function:

```{r, eval = T}
library(lavaan)
```

This is lavaan 0.6-9
lavaan is FREE software! Please report any bugs.

Don't worry about the "BETA" warning, this package is awesome! 

This may seem like a pain, but roll with it. The good news is that once you do it, you have access to a whole library of SEM functions.

If you boot up R and receive error msgs like "could not find function "sem""
IT IS PROBABLY BECAUSE YOU HAVEN'T LOADED THE lavaan PACKAGE. 

Check out the help page of a particular function, say sem():

```{r, eval = T}
help(sem)
?sem
```

<!------------------------------>
## Using Lavaan For Path Models
<!------------------------------>

Every analysis in lavaan has three main parts.

+ Part I: Writing the Model Syntax
+ Part II: Analyzing the Model Using Your Dataset
+ Part III: Examining the results.

### PART I: Follow the set of equations we wrote in class:

Self-Efficacy = BMI + Self-Esteem + Disturbance

Bulimic Symptoms = BMI + Self-Esteem + Self-Efficacy + Disturbance

Restrictive Symptoms = BMI + Self-Esteem + Self-Efficacy + Disturbance

Overall Risk = BMI + Self-Esteem + Self-Efficacy + Acculturation + Disturbance

Let's write some model syntax:

```{r, eval = T}
ex1PathSyntax <- " 			 #opening a quote
  # Tilda ~ : Regression 
  # M ~ X regression (X predicts M)
  # Each line corresponds to an equation 
  # Disturbance is automatically included for each regression 
  # (i.e. no extra term needed)
  
  DietSE ~ BMI + SelfEsteem      #DietSE is predicted by BMI and SelfEsteem 
	Bulimia ~ DietSE + BMI + SelfEsteem
	Restrictive ~ DietSE + BMI + SelfEsteem
	Risk ~ DietSE + BMI + SelfEsteem + Accu
"  
```


Things to note here:

+ We are calling our saved model syntax object ex1PathSyntax
+ We assign it using <- as usual
+ Then we open a quotation "
+ Then we write each part of the model on separate lines.
+ Then we close the quotation "
+ The variables names need to match those in the dataset (case matters!)
+ Add comments inside the model syntax using hashtag


### PART II Let's run our model!

To run this model, we will start by using the sem() function.

Sensible defaults for estimating CFA models like assumptions of linear regression, so we don’t actually have to write some constraints into the model above

Alternatively, one can use lavaan() function [with the fewest default settings] or cfa() function [with similar defaults as sem() function]

To use lavaan(), you have to specify all 22 parameters in the model. 

#### ex1fit

You can run the sem() function using two different sources of data:

1. The raw dataset, using:

```{r, eval = F}
lavaan::sem(model = modelSyntax, data = yourDataset)
```

example:

```{r, eval = F}
ex1fit <- lavaan::sem(model = ex1PathSyntax, data = labData)
```

If you encounter errors like: 

Error in if ((!is.matrix(model)) | ncol(model) != 3) stop("model argument must be a 3-column matrix") : 
  argument is of length zero

IT IS PROBABLY BECAUSE YOU HAVEN'T LOADED THE lavaan PACKAGE. 

To make sure you are using the sem() function from the lavaan package, add PackageName:: before a function: 

```{r, eval = T}
ex1fit <- lavaan::sem(model = ex1PathSyntax, data = labData)
```

Then we can obtain complete results using the summary() function:

```{r}
summary(ex1fit)
```

2. The covariance matrix, using:

```{r, eval = F}
lavaan::sem(model = modelSyntax, sample.cov = yourCovarianceMatrix, sample.nobs = numberOfObservationsInYourDataset)
```

This is to illustrate that WITH COMPLETE DATA, you can run SEM analyses using only covariances as input and obtain the same results as with raw data! 

This positions SEM for meta-analysis and replication studies. 

example:

```{r}
ex1fit_S <- lavaan::sem(model = ex1PathSyntax, sample.cov = S, sample.nobs = n)
summary(ex1fit_S)
```

The . before a variable name refers to its disturbance.

e.g., .Bulimia refers to the disturbance of Bulimia, not Bulimia itself

You should get exactly the same output in ex1fit and ex1fit_S. 

Wait, Gabriella, the df is not 6....

This is because sem() by default assumes that disturbances of endogenous variables covary among themselves (which, in our model, are not correlated at all!)

The estimates of disturbance covariances are presented under "Covariances" in the output:

```{r, eval=FALSE}
Covariances:
                   Estimate  Std.Err  z-value  P(>|z|)
.Bulimia ~~                                          
  .Restrictive       0.536    0.029   18.389    0.000
  .Risk              0.814    0.034   23.983    0.000
.Restrictive ~~                                      
  .Risk              0.785    0.034   22.996    0.000
```


#### ex1PathSyntax_noCov

To change those defaults, one needs to explicitly fix those disturbance covariances at 0 (this is a strong assumption, I know...): 

http://lavaan.ugent.be/tutorial/syntax2.html

```{r}
ex1PathSyntax_noCov <- " 			 #opening a quote
  # ~~ indicates a two-headed arrow (variance or covariance)
  # 0* in front of the 2nd variable fixes the covariance at 0
  
  DietSE ~ BMI + SelfEsteem      #DietSE is predicted by BMI and SelfEsteem 
	Bulimia ~ DietSE + BMI + SelfEsteem
	Restrictive ~ DietSE + BMI + SelfEsteem
	Risk ~ DietSE + BMI + SelfEsteem + Accu

	#Disturbance covariances (fixed at 0):
	DietSE ~~ 0*Bulimia  
	DietSE ~~ 0*Restrictive  
	DietSE ~~ 0*Risk     
	Bulimia ~~ 0*Restrictive
	Bulimia ~~ 0*Risk
	Restrictive ~~ 0*Risk
	
	# These lines above say that there is no covariance among the disturbances of all endogenous variables
"  

ex1fit_noCov <- lavaan::sem(model = ex1PathSyntax_noCov, data = labData)
summary(ex1fit_noCov)
```

```{r, eval=FALSE}
df = 6 and 

Covariances:
                  Estimate  Std.Err  z-value  P(>|z|)
.DietSE ~~                                           
  .Bulimia           0.000                           
  .Restrictive       0.000                           
  .Risk              0.000                           
.Bulimia ~~                                          
  .Restrictive       0.000                           
  .Risk              0.000                           
.Restrictive ~~                                      
  .Risk              0.000                           
```

Wait, where are the variances and covariances of exogenous variables?

They are not included in the output because they are estimated PERFECTLY

#### ex1fit_noCov_freeX

fixed.x=FALSE asks for the variances/covariances/means of the exogenous variables to be freely estimated instead of being fixed at the values found from the sample

This usually makes no difference from ex1fit_noCov, except that it prints more lines

```{r}
ex1fit_noCov_freeX <- lavaan::sem(model = ex1PathSyntax_noCov, data = labData, fixed.x = FALSE)
summary(ex1fit_noCov_freeX)
```

#### ex1fit_noCov_lavaan

As a bonus, here is how you would write the model syntax if you use lavaan() instead of sem()...

```{r}
ex1PathSyntax_lavaan <- " 			 #opening a quote
  # ~~ indicates a two-headed arrow (variance or covariance)

  #regression coefficients (12)
  DietSE ~ BMI + SelfEsteem      
	Bulimia ~ DietSE + BMI + SelfEsteem
	Restrictive ~ DietSE + BMI + SelfEsteem
	Risk ~ DietSE + BMI + SelfEsteem + Accu
	
	#variances of exogenous variables (3)
	BMI ~~ BMI
	SelfEsteem ~~ SelfEsteem
	Accu ~~ Accu

  #disturbance variances (4)
	DietSE ~~ DietSE
	Bulimia ~~ Bulimia
	Restrictive ~~ Restrictive
	Risk ~~ Risk

	#covariances among exogenous variables (3)
	BMI ~~ SelfEsteem
	BMI ~~ Accu
	SelfEsteem ~~ Accu

  #total: 22 parameters
"  
ex1fit_noCov_lavaan <- lavaan(model = ex1PathSyntax_lavaan, data = labData)
summary(ex1fit_noCov_lavaan)
```

which yields the same output as ex1fit_noCov_freeX. 

### Sigma Matrices

Let's have a look at the model-implied covarinace matrix from our final model ex1fit_noCov_freeX and save it as Sigma: 

```{r}
fitted(ex1fit_noCov_freeX)
Sigma <- fitted(ex1fit_noCov_freeX)$cov
```

How close is Sigma to S? 

+ Rearrange the rows and columns of Sigma (important!) and take the difference

```{r}
diff = Sigma[colnames(S), colnames(S)] - S
round(diff, 3)
```

How about the default model that include disturbance covariances?

```{r}
Sigma0 <- fitted(ex1fit)$cov
diff0 = Sigma0[colnames(S), colnames(S)] - S
round(diff0, 3)
```

#### Gabriella's Practical Tips:

+ To begin with, constraint the disturbance covariances to be 0 ; 
+ Keep the model if the model fits the data well;
+ Relax the constraints the disturbance covariances if the initial model did not fit well. 

### PART III: Summarizing Our Analysis:

There are some useful options we can ask for with summary():

```{r, eval = FALSE}
summary(ex1fit_noCov_freeX, fit.measures = T) #include model fit measures
summary(ex1fit_noCov_freeX, standardized = T) #This includes standardized estimates. std.all contains usual regression standardization.
summary(ex1fit_noCov_freeX, ci = T)  #Include confidence intervals

# Add them all!
```

If we JUST want the parameter estimates:
```{r}
parameterEstimates(ex1fit_noCov_freeX)
parameterEstimates(ex1fit_noCov_freeX, standardized = T) #include standardized solution....
```

For standardized solutions, there is also this function:
```{r}
standardizedSolution(ex1fit_noCov_freeX, type = "std.all") 
```

How does it work?
```{r}
?standardizedSolution
```


<!------------------------------>
## Plotting SEM model
<!------------------------------>

```{r}
# install.packages("semPlot")
library(semPlot)

# Plot!
semPaths(ex1fit_noCov_freeX)
```

```{r}
# estimates instead of paths only
semPaths(ex1fit_noCov_freeX, what='est', 
         edge.label.cex=1.25, curvePivot = TRUE, 
         fade=FALSE)
```

```{r}
# standardized solutions
semPaths(ex1fit_noCov_freeX, what='std', 
         edge.label.cex=1.25, curvePivot = TRUE, 
         fade=FALSE)
```


```{r}
semPaths(ex1fit_noCov_freeX, what='est', 
         rotation = 2, # default rotation = 1 with four options
         edge.label.cex=1.25, curvePivot = TRUE, 
         fade=FALSE)
```

### customize it your way

```{r}
semPaths(ex1fit_noCov_freeX, whatLabels="est", # plot model not parm ests
         rotation = 2, # default rotation = 1 with four options
         asize = 5, # arrows' size
         esize = 2, # width of paths' lines / curves
         edge.label.cex = 0.8, # font size of regr'n coeffs
         sizeMan = 10, # font size of manifest variable names
         nCharNodes = 0,
         nCharEdges = 0, # don't limit variable name lengths
         fade = FALSE, # don't weight path width to reflect strength
         curvePivot = TRUE, # make straight edges instead of round ones
         curve = 2, # pull covariances' curves out a little
         style = "lisrel", # no variances vs. # "ram"'s 2-headed for variances
         color = "green", # color of variables
         edge.color = "black", # color of edges/paths
         layout = "tree2", # tree, spring, circle, circle2
         residuals = TRUE) # residuals variances included in the path diagram
```

```{r}
semPaths(ex1fit_noCov_freeX, what='est', 
         rotation = 2, # default rotation = 1 with four options
         curve = 2, # pull covariances' curves out a little
         nCharNodes = 0,
         nCharEdges = 0, # don't limit variable name lengths
         sizeMan = 8, # font size of manifest variable names
         style = "lisrel", # single-headed arrows vs. # "ram"'s 2-headed for variances
         edge.label.cex=1.2, curvePivot = TRUE, 
         fade=FALSE)
```

<!------------------------------>
## Exercise: How would you fit the model in Saunders et al. (2016)?
<!------------------------------>

