# Lavaan Lab 2: Mediation and Indirect Effects

In this lab, we will learn how to: 

+ perform a simple mediation analysis using Preacher & Hayes (2004) + Bootstrap
+ test mediation effects in the eating disorder path model 

<!------------------------------>
## Reading-In and Working With Realistic Datasets In R
<!------------------------------>

If your data (eatingDisorderSimData.csv) is stored in you current working directory, then simply load your data by typing the name of the .csv file:

```{r, eval = T}
labData <- read.csv(file = "eatingDisorderSimData.csv", header = T, sep = ",")
```

<!------------------------------>
## Using Lavaan For Mediation Models - Preacher & Hayes’s 
<!------------------------------>

Load the package: 

```{r, eval = T}
library(lavaan)
```

+ Part I: Writing the Model Syntax
+ Part II: Analyzing the Model Using Your Dataset
+ Part III: Examining the results.

## PART I: # Follow the two equations of M (DietSE) & Y (Bulimia)

Diet Self-Efficacy = BMI + Disturbance

Bulimic Symptoms = BMI + Diet Self-Efficacy + Disturbance

Let's write some model syntax:

```{r, eval = T}
ex1MediationSyntax <- " 			 #opening a quote
	#Regressions
	DietSE ~ BMI                    #M ~ X regression (a path)
	Bulimia ~ BMI + DietSE          #Y ~ X + M regression (c prime and b)
	"  
```

No need to fix disturbance covariances in simple mediation as none was estimated

## PART II Let's run our model!

let fixed.x=FALSE to print more lines

```{r, eval = T}
ex1fit_freeX <- lavaan::sem(model = ex1MediationSyntax, data = labData, fixed.x = FALSE)
summary(ex1fit_freeX)
```

note that there are six parameter estimates and df = 0. 

But the output does not include the mediation effect a*b?

### Label the mediation effect

Let's learn how to label parameters

great tutorial example: http://lavaan.ugent.be/tutorial/mediation.html

To label a parameter, include the coefficient label and an asterisk * before the variable to be labelled.

E.g., y ~ b1*x + b2*m

This would give x the label b1 and m the label b2 in the y regression.

```{r, eval = T}
ex2MediationSyntax <- " 			 			#opening a quote
	#Regressions
	DietSE ~ a*BMI                  	#Label the a coefficient in the M regression.
	Bulimia ~ cPrime*BMI + b*DietSE   #Label the direct effect (cPrime) of X and direct effect of M (b) in the Y regression.
	" 
```

What does this do?

```{r}
ex2fit <- lavaan::sem(model = ex2MediationSyntax, data = labData, fixed.x=FALSE)
summary(ex2fit)
```

The regression coefficients have labels now!

### Define a new term for the mediation effect a*b 

...using the labels we just created in ex2MediationSyntax

The := operator in lavaan defines new terms to be tested:

(name of a new term) := operator

```{r}
ex3MediationSyntax <- " 			 			#opening a quote
	#Regressions
	DietSE ~ a*BMI                  	#Label the a coefficient in the M regression.
	Bulimia ~ cPrime*BMI + b*DietSE   #Label the direct effect (cPrime) of X and direct effect of M (b) in the Y regression.
	
	#Define New Parameters
	ab := a*b 									#the product term is computed as a*b
	c := cPrime + ab 						#having defined ab, we can use this here.
" 
```

```{r}
ex3fit <- lavaan::sem(model = ex3MediationSyntax, data = labData, fixed.x=FALSE)
summary(ex3fit)
```

Now there are two significance tests of the indirect effect ab and the total effect c! 

Question: why didn't the #parameters change? 

Note: *defining* a new term is NOT equivalent to *adding* a new parameter!

You can create as many terms as your want without changing the #parameters and the df


## PART III: Summarizing Our Analysis:

We can request standardized coefficients very easily by adding a statement to the summary command.

```{r}
summary(ex3fit, standardized = TRUE) #This includes standardized estimates. std.all contains usual regression standardization.

summary(ex3fit, ci = T)  #Include confidence intervals
```

or both!

```{r}
summary(ex3fit, standardized = TRUE, ci = T)
```

__Important: the default significance tests of defined parameters in lavaan is Sobel's test.__ 

## PART IV: Bootstrap confidence intervals

### The default one is boot.ci.type = "perc"

You can request bootstrap standard errors in sem() using se = "bootstrap" and bootstrap = 1000

```{r}
set.seed(2022)
ex3Boot <- lavaan::sem(model = ex3MediationSyntax, data = labData, se = "bootstrap", bootstrap = 1000, fixed.x=FALSE) 
```

This requires the full dataset - need more than the covariance matrix.

se = "bootstrap" requests bootstrap standard errors.

bootstrap = 1000 requests 1000 bootstrap samples.

Request bootstrap CI: 

```{r}
summary(ex3Boot, ci = TRUE) 
```

Now we have bootstrap standard error and percentile confidence interval for ab!

### BC (bias-corrected) confidence interval

What about other types of bootstrap confidence intervals?

You can request a BC (bias-corrected) by adding an argument boot.ci.type = "bca.simple" to parameterEstimates():

```{r}
parameterEstimates(ex3Boot, level = 0.95, boot.ci.type="bca.simple")
```

which returns a 95% BC confidence interval. 

This approach will yield similar results to the PROCESS Macro in SPSS with bias-corrected standard errors.


<!------------------------------>
## In-Class Exercise: Use Lavaan to estimate and interpret the following model
<!------------------------------>

```{r, eval=TRUE}
ex4MediationSyntax <- "
	#Regressions
	DietSE ~ a*SelfEsteem                  	
	Risk ~ cPrime*SelfEsteem + b*DietSE   
	
	#Define New Parameters
	ab := a*b 									#the product term is computed as a*b
	c := cPrime + ab 						#having defined ab, we can use this here.
"
```


```{r, eval=TRUE}
ex4fit <- lavaan::sem(model = ex4MediationSyntax, data = labData, fixed.x=FALSE)

summary(ex4fit, ci = T)
```

Bootstrap confidence intervals:

```{r, eval=TRUE}
set.seed(2022)

ex4Boot <- lavaan::sem(model = ex4MediationSyntax, data = labData, se = "bootstrap", bootstrap = 1000, fixed.x=FALSE) 

parameterEstimates(ex4Boot, level = 0.95, boot.ci.type="bca.simple", standardized = TRUE)
```


<!------------------------------>
## Exercise: Eating Disorder Mediation Analysis
<!------------------------------>

Give it a try before peaking the answers! 

Hints:

1. Label the regression coefficients: b1 - b12;

2. Fix all disturbance covariances at 0;

3. Define mediation effects and total effects for each of the six mediation models using the labels;

4. Request bootstrap standard errors using se = "bootstrap"; 

5. Print and interpret the mediation effects;

6. (Optional) Identify and interpret the inconsistent mediation effects. 

I'll get you started: 


### Step 1: Labeling and defining the parameters
### Step 2: Fix all disturbance covariances at 0

```{r}
ex5PathSyntax_noCov <- " 			 #opening a quote
	DietSE ~ b1*BMI + b5*SelfEsteem      #DietSE is predicted by BMI and SelfEsteem 
	Bulimia ~ b10*DietSE + b2*BMI + b6*SelfEsteem
	Restrictive ~ b11*DietSE + b3*BMI + b7*SelfEsteem
	Risk ~ b12*DietSE + b4*BMI + b8*SelfEsteem + b9*Accu

	#Disturbance covariances (fixed at 0):
	DietSE ~~ 0*Bulimia  # ~~ indicates a two-headed arrow (variance or covariance)
	DietSE ~~ 0*Restrictive  # 0* in front of the 2nd variable fixes the covariance at 0
	DietSE ~~ 0*Risk     # These lines say that all endogenous variables have no correlated disturbance variances
	Bulimia ~~ 0*Restrictive
	Bulimia ~~ 0*Risk
	Restrictive ~~ 0*Risk
"  
```


### Step 3: Define new terms for mediation effects

Recall:

Define New Parameters

ab := a*b 									#the product term is computed as a*b


```{r, eval=TRUE}
ex5MediationSyntax <- "
	DietSE ~ b1*BMI + b5*SelfEsteem      #DietSE is predicted by BMI and SelfEsteem 
	Bulimia ~ b10*DietSE + b2*BMI + b6*SelfEsteem
	Restrictive ~ b11*DietSE + b3*BMI + b7*SelfEsteem
	Risk ~ b12*DietSE + b4*BMI + b8*SelfEsteem + b9*Accu

	#Disturbance covariances (fixed at 0):
	DietSE ~~ 0*Bulimia  # ~~ indicates a two-headed arrow (variance or covariance)
	DietSE ~~ 0*Restrictive  # 0* in front of the 2nd variable fixes the covariance at 0
	DietSE ~~ 0*Risk     # These lines say that all endogenous variables have no correlated disturbance variances
	Bulimia ~~ 0*Restrictive
	Bulimia ~~ 0*Risk
	Restrictive ~~ 0*Risk
	
	#Define New Parameters
	med1 := b1*b10
	total1 := b2 + med1
	med2 := b1*b11
	total2 := b3 + med2
	med3 := b1*b12
	total3 := b4 + med3
	med4 := b5*b10
	total4 := b6 + med4
	med5 := b5*b11
	total5 := b7 + med5
	med6 := b5*b12
	total6 := b8 + med6
	
	# difference term for significance testing 
	diff1 := med1 - med4
"
```


```{r, eval=FALSE}
ex5fit <- lavaan::sem(model = ex5MediationSyntax, data = labData, fixed.x=FALSE)

summary(ex5fit, ci = T)
```

### Step 4: Bootstrap confidence intervals:
### Step 5: Print and interpret the mediation effects;

```{r, eval=TRUE}
set.seed(2022)

ex5Boot <- lavaan::sem(model = ex5MediationSyntax, data = labData, se = "bootstrap", bootstrap = 1000, fixed.x=FALSE) 

parameterEstimates(ex5Boot, level = 0.95, boot.ci.type="bca.simple", standardized = TRUE)
```


### Plot it!

```{r}
library(semPlot)

semPaths(ex5Boot, what='est', 
         rotation = 2, # default rotation = 1 with four options
         curve = 2, # pull covariances' curves out a little
         nCharNodes = 0,
         nCharEdges = 0, # don't limit variable name lengths
         sizeMan = 8, # font size of manifest variable names
         style = "lisrel", # single-headed arrows vs. # "ram"'s 2-headed for variances
         edge.label.cex=1.2, curvePivot = TRUE, 
         fade=FALSE)
```

