# Lavaan Lab 5: One-factor CFA Model

In this lab, we will learn how to: 

+ Identify the One-factor CFA Model
+ Scale the One-factor CFA Model
+ Estimate the One-factor CFA Model
+ Interpret the One-factor CFA Model


## Data Prep

We will use cfaInClassData.csv in this lab. 

This is a simulated dataset based on Todd Little's positive affect example.

The hypothesis is that a latent variable ‘positive affect’ is measured by three indicators (glad, cheerful, and happy). 

Let's read this dataset in: 

```{r}
cfaData<- read.csv("cfaInclassData.csv", header = T)
```

and examine the dataset: 

```{r}
head(cfaData)
str(cfaData)
dim(cfaData) #n = 1000, 7 variables
```

Let's examine their means and standard deviations:

```{r}
round(apply(cfaData[,-1], 2, mean), 2) # mean-centered
round(apply(cfaData[,-1], 2, sd), 2) 
```

Let's call up the lavaan library and run some CFA's!

```{r}
library(lavaan)
```

<!------------------------------>
## PART I: One-Factor CFA, Fixed Loading
<!------------------------------>

### Fixed Loading, AKA Marker Variable method. 

FYI, the three equations for the three indicators are:

+ Glad = lambda1*posAffect(eta) + u1
+ Cheerful = lambda2*posAffect(eta) + u2
+ Happy = lambda3*posAffect(eta) + u3

Let's first follow the equations above and write the syntax (disturbances are automatically included): 

```{r, eval=FALSE}
mod1.wrong<- "
  glad ~ posAffect
  cheerful ~ posAffect
  happy ~ posAffect
"
fit1.wrong = lavaan::sem(model = mod1.wrong, data = cfaData, fixed.x=FALSE)
```

Oops - an error message! 

```{r,eval=FALSE}
Error in lav_data_full(data = data, group = group, cluster = cluster,  : 
  lavaan ERROR: missing observed variables in dataset: posAffect
```

This is because posAffect is a latent variable and we have to use =~ to define a latent variable: 

```{r, eval=FALSE}
mod1.wrong<-'
posAffect =~ Glad + Cheerful + Happy
'
fit1.wrong = lavaan::sem(model = mod1.wrong, data = cfaData, fixed.x=FALSE)
```

```{r,eval=FALSE}
Error in lavaan::lavaan(model = mod1.wrong, data = cfaData, fixed.x = FALSE,  : 
  lavaan ERROR: missing observed variables in dataset: Glad Cheerful Happy
```

Error, why? 

The variable names in the model syntax have to match the column names EXACTLY, even the letter cases. 

Let's try again: 

```{r}
mod1<-'
posAffect =~ glad + cheerful + happy
'
```

Let's explain the lavaan model syntax!

+ mod1 is used to name our model. 
+ Since posAffect is a latent variable (it's not in the data), we cannot follow the equations above and write syntax like glad ~ posAffect
+ Instead, we specify a CFA measurement model in mod1.
+ NEW SYNTAX ALERT: Using =~ means "manifested by"
+ In the code above we can see that our latent construct 'posAffect' is manifested by glad, cheerful, and happy
+ By default, the loading of glad is fixed at 1 (Fixed Loading Method)

Next we name the fitted object 'fit1' to see our output. 

```{r}
fit1 = lavaan::sem(mod1, data = cfaData, fixed.x=FALSE)
```

This summary will show us the loadings (I also requested standardized results):

```{r}
summary(fit1, standardized = T)
```

```{r, eval=FALSE}
df = 0 (why?)

Latent Variables:
                   Estimate  Std.Err  z-value  P(>|z|)   Std.lv  Std.all
  posAffect =~                                                          
    glad              1.000                               0.693    0.705
    cheerful          1.117    0.059   18.782    0.000    0.774    0.787
    happy             1.066    0.057   18.786    0.000    0.739    0.757
```

What does this mean?

+ 1 unit change in posAffect produces:
  * 1-unit change in "glad" (marker indicator)
  * 1.117-unit change in "cheerful" (1.117 times greater than the effect on "glad")
  * 1.066-unit change in "happy" (1.066 times greater than the effect on "glad")

```{r, eval=FALSE}
Variances:
Unique factor variances:
                   Estimate  Std.Err  z-value  P(>|z|)   Std.lv  Std.all
   .glad              0.485    0.030   16.238    0.000    0.485    0.503
   .cheerful          0.367    0.030   12.062    0.000    0.367    0.380
   .happy             0.407    0.030   13.751    0.000    0.407    0.427
```

+ The leftover unique factor variances remain substantial
+ Meaning that none of the indicators is a perfect measure of posAffect
+ but they all contribute significantly to the measurement of posAffect (the standardized loadings above larger than 0.6)

Followed by the latent factor variance.
```{r, eval=FALSE}
                   Estimate  Std.Err  z-value  P(>|z|)   Std.lv  Std.all
    posAffect         0.480    0.043   11.270    0.000    1.000    1.000
```

### Change marker indicator

If you'd like to fix the 2nd loading to 1:

```{r}
mod1b_wrong<-'
posAffect =~ glad + 1*cheerful + happy
'
```

won't work. 

You will get something like this:

```{r,eval=FALSE}
Latent Variables:
                   Estimate  Std.Err  z-value  P(>|z|)   Std.lv  Std.all
  posAffect =~                                                          
    glad              1.000                               0.734    0.733
    cheerful          1.000                               0.734    0.759
    happy             1.009    0.046   22.052    0.000    0.741    0.759
```

You'll have to change the order of the indicators to move cheerful to the front of the variable list: 

```{r}
mod1b<-'
posAffect =~ cheerful + glad + happy
'
```

Or use \*NA to specify which loading to keep free and use *1 to specify the marker variable whose loading to be fixed at 1

```{r}
mod1b<-'
posAffect =~ NA*glad + 1*cheerful + NA*happy
'
```

Here we named the fitted object 'fit1b' to see our output. 

```{r}
fit1b = lavaan::sem(mod1b, data = cfaData, fixed.x=FALSE)
summary(fit1b, standardized = T)
```

+ The loadings can be obtained by dividing those in fit1 by 1.117 (i.e., they change proportionally).
+ The variances of unique factors and latent factor remain unchanged. 

<!------------------------------>
## PART II: One-Factor CFA, Fixed Factor Variance
<!------------------------------>

### Fixed Factor Method

Keep using the same syntax but assign a new name mod2: 

```{r}
mod2<-'
posAffect =~ glad + cheerful + happy
'
```

To fix the variance of the latent variable to 1, add std.lv=T to sem() function: 

```{r}
fit2<-lavaan::sem(mod2, data = cfaData, fixed.x=FALSE, std.lv=T)
summary(fit2, standardized = TRUE)
```

```{r, eval=FALSE}
Latent Variables:
                   Estimate  Std.Err  z-value  P(>|z|)   Std.lv  Std.all
  posAffect =~                                                          
    glad              0.693    0.031   22.540    0.000    0.693    0.705
    cheerful          0.774    0.031   25.233    0.000    0.774    0.787
    happy             0.739    0.030   24.226    0.000    0.739    0.757
```

+ 1-SD change in the factor (posAffect) causes:
  * 0.693-unit change in glad (on its raw scale)
  * 0.774-unit change in cheerful (on its raw scale)
  * 0.739-unit change in happy (on its raw scale)

```{r, eval=FALSE}
Variances:
                   Estimate  Std.Err  z-value  P(>|z|)   Std.lv  Std.all
   .glad              0.485    0.030   16.238    0.000    0.485    0.503
   .cheerful          0.367    0.030   12.062    0.000    0.367    0.380
   .happy             0.407    0.030   13.751    0.000    0.407    0.427
    posAffect         1.000                               1.000    1.000
```

+ We see that posAffect now has variance (=sd) of 1
+ All loadings were freely estimated, no loading is 1.
+ and the unique factor variances are the same as before

<!------------------------------>
## Exercise: One-factor CFA Model
<!------------------------------>

Could you use the indicators satisfied, content, and comfortable to build a one-factor CFA model to measure a latent variable called Satisfaction? 

Use the Fixed Loading and the Fixed Factor Methods and compare their estimates. 

### Fixed Loading

```{r}

```

### Fixed Factor

```{r}

```

