# Lavaan Lab 6: Two-factor CFA Model


## Data Prep

We will continue to use cfaInClassData.csv in this lab. 

Let's read this dataset in: 

```{r}
cfaData<- read.csv("cfaInclassData.csv", header = T)
```

Load up the lavaan library and run some CFA's!

```{r}
library(lavaan)
```

<!------------------------------>
## PART I: Two-Factor CFA, Fixed Loading
<!------------------------------>

### Fixed Loading, AKA Marker Variable method. 

Let's write up the model syntax for the measurement model with two factors:

```{r}
fixedIndTwoFacSyntax <- "
	#Factor Specification	
	posAffect    =~    glad + happy + cheerful 
	satisfaction =~ satisfied + content + comfortable
"
```

Here we named the fitted object 'fixedIndTwoFacRun' to see our output: 

```{r}
fixedIndTwoFacRun = lavaan::sem(model = fixedIndTwoFacSyntax, data = cfaData, fixed.x=FALSE)
```

Get a summary using summary() function, add standardized=T to request standardized parameter estimates: 

```{r}
summary(fixedIndTwoFacRun, standardized=T)
```

```{r}
summary(fixedIndTwoFacRun, standardized = T)
```

Here is the fun part. Plot the fitted model using semPaths() function from the semPlot package: 

```{r}
library(semPlot)
semPaths(fixedIndTwoFacRun)
semPaths(fixedIndTwoFacRun, what = 'est') # under "Estimate"
```

```{r, eval=FALSE}
df = 8 (why?)

Latent Variables:
                     Estimate  Std.Err  z-value  P(>|z|)   Std.lv  Std.all
  posAffect =~                                        
     glad              1.000                               0.694    0.706
     happy             1.067    0.055   19.294    0.000    0.740    0.758
     cheerful          1.112    0.057   19.458    0.000    0.772    0.785
   satisfaction =~                                                       
     satisfied         1.000                               0.773    0.767
     content           1.068    0.052   20.525    0.000    0.826    0.762
     comfortable       0.918    0.045   20.336    0.000    0.709    0.746
```

What does this mean?

+ 1 unit change in posAffect (factor) produces:
  * 1-unit change in "glad" (marker indicator)
  * 1.067-unit change in "happy" (1.067 times greater than the effect on "glad")
  * 1.112-unit change in "cheerful" (1.112 times greater than the effect on "glad")

+ 1 unit change in satisfaction (factor) produces:
  * 1-unit change in "satisfied" (marker indicator)
  * 1.068-unit change in "content" (1.068 times greater than the effect on "satisfied")
  * 0.918-unit change in "comfortable" (0.918 times greater than the effect on "satisfied")

```{r, eval=FALSE}
Variances:
                    Estimate  Std.Err  z-value  P(>|z|)   Std.lv  Std.all
    .glad              0.484    0.029   16.647    0.000    0.484    0.501
    .happy             0.405    0.028   14.389    0.000    0.405    0.425
    .cheerful          0.371    0.029   13.004    0.000    0.371    0.384
    .satisfied         0.419    0.029   14.326    0.000    0.419    0.412
    .content           0.491    0.034   14.542    0.000    0.491    0.419
    .comfortable       0.400    0.026   15.315    0.000    0.400    0.443
```

+ The leftover unique factor variances remain substantial
+ Meaning that none of the indicators is a perfect measure of posAffect or satisfaction
+ but they all contribute significantly to the measurement of latent variables (the standardized loadings above larger than 0.6)

Followed by two factor variances:
```{r, eval=FALSE}
                    Estimate  Std.Err  z-value  P(>|z|)   Std.lv  Std.all
     posAffect         0.482    0.042   11.439    0.000    1.000    1.000
     satisfaction      0.597    0.047   12.686    0.000    1.000    1.000
```

+ which were freely estimated using Fixed Loading scaling method. 
+ Both posAffect and satisfaction are variable across participants (sig* according to p-values)
+ posAffect seems to be more stable than satisfaction (0.482<0.597). 

```{r, eval=FALSE}
Covariances:
                  Estimate  Std.Err  z-value  P(>|z|)   Std.lv  Std.all
 posAffect ~~                                                          
   satisfaction      0.262    0.025   10.284    0.000    0.488    0.488
```

+ covariance between posAffect and satisfaction is 0.262
+ standardized covariance (correlation) between posAffect and satisfaction is 0.488 (sig*)
+ 0.262/sqrt(0.482)/sqrt(0.597) = 0.488
+ posAffect is positively correlated with satisfaction. Participants who have a high level posAffect tend to have a high level of satisfaction. 

### How well does this model fit to the data? 

The model-implied variance of glad is the sum of:

+ Tracing 1: lam1\*psi1\*lam1 = 1.0\*0.482\*1.0 = 0.482
+ Tracing 2: sig2_u1 = sig2_glad = 0.484

which is 0.966. 

What is the sample variance of glad? 

+ var(cfaData$glad) = 0.9664733, which is super close to the model-implied one! 
+ meaning our CFA model is doing a pretty good job at explaining the sample variance of glad. 

The proportion of variance of glad explained by posAffect is: 

+ Tracing 1 / (Tracing 1+Tracing 2) = 49.9%
+ Unexplained: 50.1%

[Exercise] What about the model-implied variance of comfortable? 

+ Tracing 1: 0.918\*.597\*.918 = .503
+ Tracing 2: .400
+ Total .903
+ Proportion of variance explained: .503/.903
+ Sample var: var(cfaData$comfortable) = 0.904

[Exercise] What about the model-implied covariance between glad and comfortable? (e.g., cov(y1, y6))

+ Tracing 1: 1\*.262\*.918 = .241
+ Sample cov: cov(cfaData$glad, cfaData$comfortable) = .230

Overall speaking, how is our model doing at explaining the sample covariance matrix? 
Remember the fitted() function we used during path analysis R labs?

Apply fitted() function to the fitted object fixedIndTwoFacRun (not on the model syntax fixedIndTwoFacSyntax!!):

```{r}
fitted(fixedIndTwoFacRun)
Sigma <- fitted(fixedIndTwoFacRun)$cov
```

fixed variance: 

```{r}
fixedVarTwoFacRun = lavaan::sem(model = fixedIndTwoFacSyntax, data = cfaData, fixed.x=FALSE, std.lv = TRUE)
Sigma2 <- fitted(fixedVarTwoFacRun)$cov
Sigma2 # same as Sigma!
```


How close is Sigma to S? 

+ Rearrange the rows and columns of Sigma (important!) and take the difference between Sigma and S

```{r}
(S = cov(cfaData[,-1]))

diff = Sigma[colnames(S), colnames(S)] - S
round(diff, 4)

print(paste0("The difference between S and Sigma ranged between ", round(min(diff),4), " and ", round(max(diff),4), "."))
```

This is the closest Sigma can get to S. Any other set of parameter estimates would yield bigger differences with S. 

Have you wondered about how we obtain the parameter estimates in the output?

Estimation down the road...

### Fundamental Equation of SEM

Just some bonus stuff...

You can inspect the fitted object using inspect() and save the object as InspFit1. 

```{r}
InspFit1 <- inspect(fixedIndTwoFacRun, what = "est")
```

Extract Lambda matrix from InspFit1. 

```{r}
Lambda <- InspFit1$lambda
Lambda
```

Extract Psi matrix from InspFit1. 

```{r}
Psi <- InspFit1$psi
Psi
```

Extract Theta matrix from InspFit1. 

```{r}
Theta <- InspFit1$theta
Theta
```

Use the three matrices above to calculate the model-implied covariance matrix and save is as SIGMA: 

```{r}
SIGMA <- Lambda%*%Psi%*%t(Lambda)+Theta
SIGMA
```

A shortcut function to obtain the SIGMA matrix is to use fitted() function, as shown above...

```{r}
all.equal(Sigma, SIGMA)
```

### Interpretion

How do I interpret the results? 

+ (1) Introduce the scaling method I used;
+ (2) Based on the loadings, acknowledge that the indicators are not perfect measures of the latent factors, but they all contribute significantly to the measurement of latent factors (the standardized loadings above larger than 0.6);
+ (3) Report the the proportions of variance explained on each indicator;
+ (4) Describe the discrepancy between S and Sigma (where the main differences lie, and whether the differences are concerning);
+ (5) Interpret the correlation among the latent factors (size, sign, positive/negative, sig/non-sig). 


### Standardized solutions: Std.lv vs. Std.all

Std.lv: 

+ This is the solution you'll get using Fixed Factor Variance Scaling Method;
+ All factor variances are fixed as 1.0;
+ Factor covariance is the same as factor correlation;
+ All factor loadings are freely estimated so that the model-implied covariance matrix remains the same;
+ All unique factor variances remain unchanged. 

For example, under Std.lv, the model-implied variance of glad is the sum of:

+ Tracing 1: lam1\*psi1\*lam1 = 0.694\*1.0\*0.694 = 0.482
+ Tracing 2: sig2_u1 = sig2_glad = 0.484
+ Proportion of variance explained: 0.482/0.966 = 50.1%

which is still 0.966. 

[Exercise] Why is 0.709 the loading of comfortable under Std.lv? 

Under Std.lv, the model-implied variance of comfortable is the sum of:

+ Tracing 1: 
+ Tracing 2: 
+ Proportion of variance explained: 


Std.all:

+ All factor variances are fixed as 1.0;
+ Factor covariance is the same as factor correlation;
+ All factor loadings and unique factor variances are re-estimated so that the model-implied variances of indicators are all 1.0;

For example, under Std.all, the model-implied variance of glad is the sum of:

+ Tracing 1: lam1\*psi1\*lam1 = 0.706\*1.0\*0.706 = 0.499
+ Tracing 2: sig2_u1 = sig2_glad = 0.501

which add to 1.0. 

[Exercise] Why is 0.746 the loading of comfortable under Std.all? 

Under Std.all, the model-implied variance of comfortable is the sum of:

+ Tracing 1: 
+ Tracing 2: 


Why the hassle? 

+ The solution is fully standardized so that squaring Std.all loadings is equivalent to the proportion of variance explained: 
+ 0.706\*0.706 (std.all)
+ 1.0\*0.482\*1.0/0.966 (fixed loading)
+ 0.694\*1.0\*0.694/0.966 (fixed variance; std.lv)
+ = 49.9%


<!------------------------------>
## PART II: Two-Factor CFA, Fixed Factor Variance
<!------------------------------>

### Fixed Factor Method

Keep using the same syntax but assign a new name: 

```{r}
fixedFacTwoFacSyntax <- "
	#Factor Specification	
	posAffect =~ glad + happy + cheerful 
	satisfaction =~ satisfied + content + comfortable
"
```

To fix the variance of the latent variable to 1, add std.lv=T to sem() function: 

```{r}
fixedFacTwoFacRun = lavaan::sem(model = fixedFacTwoFacSyntax, 
                        data = cfaData, 
                        fixed.x=FALSE, 
                        std.lv=T)
```

Get a summary using summary() function, add standardized=T to request standardized parameter estimates 

```{r}
summary(fixedFacTwoFacRun, standardized=T)
```


```{r, eval=FALSE}
 df = 8  # same!

Latent Variables:
                   Estimate  Std.Err  z-value  P(>|z|)   Std.lv  Std.all
  posAffect =~                                                          
    glad              0.694    0.030   22.878    0.000    0.694    0.706
    happy             0.740    0.030   24.806    0.000    0.740    0.758
    cheerful          0.772    0.030   25.798    0.000    0.772    0.785
  satisfaction =~                                                       
    satisfied         0.773    0.030   25.373    0.000    0.773    0.767
    content           0.826    0.033   25.207    0.000    0.826    0.762
    comfortable       0.709    0.029   24.584    0.000    0.709    0.746
```

+ 1-SD change in the factor (posAffect) causes:
  * 0.694-unit change in glad (on its raw scale)
  * 0.740-unit change in happy (on its raw scale)
  * 0.772-unit change in cheerful (on its raw scale)

+ 1-SD change in the factor (satisfaction) causes:
  * 0.773-unit change in satisfied (on its raw scale)
  * 0.826-unit change in content (on its raw scale)
  * 0.709-unit change in comfortable (on its raw scale)

```{r, eval=FALSE}
 Covariances:
                    Estimate  Std.Err  z-value  P(>|z|)   Std.lv  Std.all
   posAffect ~~                                                          
     satisfaction      0.262    0.025   10.284    0.000    0.488    0.488

Variances:
                    Estimate  Std.Err  z-value  P(>|z|)   Std.lv  Std.all
    .glad              0.484    0.029   16.647    0.000    0.484    0.501
    .happy             0.405    0.028   14.389    0.000    0.405    0.425
    .cheerful          0.371    0.029   13.004    0.000    0.371    0.384
    .satisfied         0.419    0.029   14.326    0.000    0.419    0.412
    .content           0.491    0.034   14.542    0.000    0.491    0.419
    .comfortable       0.400    0.026   15.315    0.000    0.400    0.443
```

remain unchanged. 

Followed by two factor variances.

```{r, eval=FALSE}
                    Estimate  Std.Err  z-value  P(>|z|)   Std.lv  Std.all
    posAffect         1.000                               1.000    1.000
    satisfaction      1.000                               1.000    1.000
```

