# Lavaan Lab 7: Two-factor SR Model


## Data Prep

Again, we use cfaInClassData.csv in this lab 

Let's read this dataset in: 

```{r}
cfaData<- read.csv("cfaInclassData.csv", header = T)
```

Load up the lavaan library: 

```{r}
library(lavaan)
```

<!------------------------------>
## PART I: Two-Factor SR, Fixed Loading
<!------------------------------>

### Fixed Loading, AKA Marker Variable method. 

Let's write up the model syntax for the structural regression (SR) model with two factors:

```{r}
srSyntax <- "
	#Factor Specification	
	posAffect =~ glad + cheerful + happy  
	satisfaction =~ satisfied + content + comfortable 
	
	#Structural Regression!
	satisfaction ~ posAffect
"
```

Here we named the fitted object 'srRun' to see our output: 

```{r}
srRun = lavaan::sem(model = srSyntax, 
                    data = cfaData, 
                    fixed.x=FALSE)
```

Get a summary using summary() function, add standardized=T to request standardized parameter estimates:

```{r}
summary(srRun, standardized = T)
```

The above syntax reproduces the SR analysis from the class slides.

+ CFA part under "Latent Variables" remains unchanged
+ "Covariances" section no longer exists
+ satisfaction is also removed under "Variances"

Instead, in the Regressions section: 

```{r,eval=FALSE}
Regressions:
                   Estimate  Std.Err  z-value  P(>|z|)   Std.lv  Std.all
  satisfaction ~                                                        
    posAffect         0.544    0.047   11.490    0.000    0.488    0.488
```

returns the regression slope of posAffect (b = 0.544). One-unit change in posAffect is leading to one-unit change in satisfaction. 

```{r,eval=FALSE}
Variances:
                   Estimate  Std.Err  z-value  P(>|z|)   Std.lv  Std.all
   .satisfaction      0.455    0.038   11.869    0.000    0.762    0.762
```

returns the disturbance variance of satisfaction (sigma_(d_2)^2, not psi_2!)


<!------------------------------>
## PART II: Two-Factor SR, Fixed Factor Variance
<!------------------------------>

### Fixed Factor Method

Here we named the fitted object 'srRun2'. 

We add std.lv=T to fix all latent factor variances to 1: 

```{r}
srRun2 = lavaan::sem(model = srSyntax, 
            data = cfaData, 
            fixed.x=FALSE,
            std.lv=T)
summary(srRun2, standardized = T)
```

This doesn't work as expected, why?

Therefore, for SR models, we should always go with fixed loading scaling approach. 



<!------------------------------>
## PART III: Exercise (what fun!): 3-Factor SR Model
<!------------------------------>

Suppose that in a 3-factor SR model: 

+ Positive Affect is measured by Happy and Cheerful
+ Satisfaction is measured by Satisfied and Content
+ Pleasure is measured by Glad and Comfortable
+ Satisfaction is predicted by both Positive Affect and Pleasure

Can you use cfaData to fit such a model? 

```{r}

```

