# Lavaan Lab 8: Estimation Methods


In this lab, we will learn how to estimate parameters in CFA/SR models. 

Load up the lavaan library: 

```{r}
library(lavaan)
```

<!------------------------------>
## PART I: Hypothetical Example
<!------------------------------>

### One-factor CFA model

A made-up sample covariance matrix with n = 200:

```{r}
n = 200
S_3fac = matrix(c(5, 2, 3.5, 2, 3, 2, 3.5, 2, 6), 3, 3, 
                dimnames = list(c('Y1', 'Y2', 'Y3'), c('Y1', 'Y2', 'Y3')))
S_3fac
```
Fit a one-factor CFA to the sample covariance matrix:

```{r}
one_fac_syntax <- "
	eta =~ Y1 + Y2 + Y3
"
```

Request Unweighted Least Squares (ULS):

```{r}
one_fac_fit2 <- lavaan::sem(one_fac_syntax, 
                    sample.cov = S_3fac, 
                    sample.nobs = n, 
                    estimator = "ULS", 
                    fixed.x = FALSE)

summary(one_fac_fit2, standardized = T)
```

Sigma:

```{r}
fitted(one_fac_fit2)$cov
```

```{r}
Sigma = fitted(one_fac_fit2)$cov
diff = Sigma[colnames(S_3fac), colnames(S_3fac)] - S_3fac
round(diff,3)
```
all zeros. Meaning that Sigma = S.


<!------------------------------>
## PART II: ULS on the Positive Affect Example
<!------------------------------>

Let's read this dataset in: 

```{r}
cfaData<- read.csv("cfaInclassData.csv", header = T)
```

Fit a two-factor CFA model: 

```{r}
fixedIndTwoFacSyntax <- "
	#Factor Specification	
	posAffect =~ glad + happy + cheerful 
	satisfaction =~ satisfied + content + comfortable
"
```

```{r}
two_fac_fit_uls <- lavaan::sem(fixedIndTwoFacSyntax, 
                       data = cfaData, 
                       fixed.x = FALSE,
                       estimator = "ULS")

summary(two_fac_fit_uls, standardized = T)
```

Sigma:

```{r}
S = cov(cfaData[,-1])
Sigma = fitted(two_fac_fit_uls)$cov[colnames(S), colnames(S)]
diff = Sigma - S
round(diff,3)

print(paste0("The difference between S and Sigma ranged between ", round(min(diff),4), " and ", round(max(diff),4), "."))
```
Sigma is not the same as S, but close. 

The sum of squared differences is:

```{r}
sum(diff[lower.tri(diff,diag = T)]^2)
```



<!------------------------------>
## PART III: Calculate ULS test statistic manually
<!------------------------------>

ULS test statistic is calculated as:

```{r}
T_uls = (1000-1)*sum(diff[lower.tri(diff,diag = T)]^2)
T_uls
```

One can also obtain vectors of S and Sigma first: 

```{r}
s = lav_matrix_vech(S)
sigma = lav_matrix_vech(Sigma)
```

and calculate the ULS test statistic: 

```{r}
T_uls = (1000-1)*sum((s - sigma)^2)
T_uls
```
+ No p-value for ULS test statistic
+ as there is no known distribution for this test statistic
+ i.e., no suitable method for model fit evaluation



<!------------------------------>
## PART IV: ML vs ULS vs WLS
<!------------------------------>

### ML Estimation

```{r}
two_fac_fit_ml <- lavaan::sem(fixedIndTwoFacSyntax, 
                       data = cfaData, 
                       fixed.x = FALSE,
                       estimator = "ML")
```

### WLS Estimation

```{r}
two_fac_fit_wls <- lavaan::sem(fixedIndTwoFacSyntax, 
                      data = cfaData, 
                      fixed.x = FALSE,
                      estimator = "WLS")
```

### Compare the parameter estimates

```{r}
coefTable = parameterEstimates(two_fac_fit_ml)[,1:3]
coefTable = cbind(coefTable, 
                  ML = parameterEstimates(two_fac_fit_ml)$est, 
                  ULS = parameterEstimates(two_fac_fit_uls)$est, 
                  WLS = parameterEstimates(two_fac_fit_wls)$est)
coefTable
```

### Compare the standard errors:

```{r}
seTable = parameterEstimates(two_fac_fit_ml)[,1:3]
seTable = cbind(seTable, 
                  ML = parameterEstimates(two_fac_fit_ml)$se, 
                  ULS = parameterEstimates(two_fac_fit_uls)$se, 
                  WLS = parameterEstimates(two_fac_fit_wls)$se)
seTable
```


<!------------------------------>
## PART V: Improper Solutions
<!------------------------------>

Going back to the 1-factor toy example...

Suppose we have a new covariance matrix now and a super large sample size (n = 200000000): 

```{r}
n = 2000
S_3fac_new = matrix(c(5, 1, 3.5, 1, 3, 2, 3.5, 2, 6), 3, 3, 
                    dimnames = list(c('Y1', 'Y2', 'Y3'), c('Y1', 'Y2', 'Y3')))
S_3fac_new
```

The one-factor syntax is the same:

```{r}
one_fac_syntax <- "
	eta =~ Y1 + Y2 + Y3
"
```

ML Estimation:

```{r}
one_fac_fit_new <- lavaan::sem(one_fac_syntax, 
                    sample.cov = S_3fac_new, 
                    sample.nobs = n, 
                    estimator = "ML", 
                    fixed.x = FALSE)
```
+ lavaan WARNING: some estimated ov variances are negative
+ negative residual variances
+ changing the estimation from ML to ULS doesn't help

```{r}
summary(one_fac_fit_new, standardized = T)
```

Label and constraint sig3 to be larger than 0:

```{r}
one_fac_syntax_const <- "
	eta =~ Y1 + Y2 + Y3
	
	Y3~~sig3*Y3
	# constraints
	sig3 > 0
"
```

ML Estimation:

```{r}
one_fac_fit_new2 <- lavaan::sem(one_fac_syntax_const, 
                       sample.cov = S_3fac_new, 
                       sample.nobs = n, 
                       fixed.x = FALSE)
summary(one_fac_fit_new2, standardized = T)
```

```{r}
fitted(one_fac_fit_new)$cov # perfect
fitted(one_fac_fit_new2)$cov # compromised, at the cost of model fit 
S_3fac_new
```

