# Lavaan Lab 9: Model Fit Part I (Test Statistics)


In this lab, we will learn: 

+ how to calculate and interpret chi-square statistics for SEM models.
+ how to compare nested models using chi-square difference test. 

Load up the lavaan and semPlot libraries: 

```{r}
library(lavaan)
library(semPlot)
```

<!------------------------------>
## PART I: Robust ML on the Positive Affect Example
<!------------------------------>

Let's read this dataset in: 

```{r}
cfaData<- read.csv("cfaInclassData.csv", header = T)
```

Write out syntax for a two-factor CFA model: 

```{r}
fixedIndTwoFacSyntax <- "
	#Factor Specification	
	posAffect =~ glad + happy + cheerful 
	satisfaction =~ satisfied + content + comfortable
"
```

Fit the model regularly:

```{r}
fixedIndTwoFacRun = lavaan::sem(model = fixedIndTwoFacSyntax, 
                        data = cfaData, 
                        fixed.x=FALSE)

fixedIndTwoFacRun
```
+ Model chi_sq: misfit defined through the likelihood ratio.
+ T_ML = 2.957
+ df_ML = 8
+ pvalue_ML = 0.937
+ Since pvalue_ML > 0.05, this model does not have significant model misfit. 

### Mean corrected statistic (T_M)

Satorra and Bentler (1994, 2001) proposed two robust corrections for non-normally distributed data: 

```{r}
two_fac_fit_M <- lavaan::sem(fixedIndTwoFacSyntax, 
                     data = cfaData, 
                     fixed.x=FALSE,
                     estimator = "MLM")

two_fac_fit_M
```

+ T_M = 2.891
+ df_M = 8
+ pvalue_M = 0.941

### Mean and variance adjusted statistic (T_MV)

+ estimator = "MLMVS" returns the Mean- and variance adjusted statistic with an updated degrees of freedom (recommended)
+ estimator = "MLMV" returns another version of Mean- and variance adjusted statistic but does not change the degrees of freedom 

```{r}
two_fac_fit_MV <- lavaan::sem(fixedIndTwoFacSyntax, 
                      data = cfaData, 
                      fixed.x=FALSE,
                      estimator = "MLMVS")

#summary(two_fac_fit_MV, standardized = T)
two_fac_fit_MV
```

+ T_MV = 2.842
+ df_MV = 7.864
+ pvalue_MV = 0.939

Please see a complete list of estimators here: http://lavaan.ugent.be/tutorial/est.html

### Yuan-Bentler test statistic (T_MLR)

Just like T_M and T_MV, T_MLR also corrects for nonnormality. Since MLR works for both complete and incomplete data, T_MLR is more popular in practice:

```{r}
two_fac_fit_MLR <- lavaan::sem(fixedIndTwoFacSyntax, 
                      data = cfaData, 
                      fixed.x=FALSE,
                      estimator = "MLR")

two_fac_fit_MLR
```
+ T_MLR = 2.897
+ df_MLR = 8
+ pvalue_MLR = 0.941

### Small sample correction - F test

```{r}
F_ratio = 2.957/8
```

referred to an F(df, N-1) distribution:

```{r}
?pf
p_val_F = 1-pf(F_ratio, 8, 1000-1)
p_val_F
```

which is very similar to the three p-values given above given a large sample size in this study N = 1000. 

<!------------------------------>
## PART II: Nested Model Comparison
<!------------------------------>

We can compare the fit of the two-factor model to that of a one-factor model because the one-factor model is nested in the two-factor model. 

### One-factor model

```{r}
OneFacSyntax <- "
	#Factor Specification	
	eta1 =~ glad + happy + cheerful + satisfied + content + comfortable  
"

one_fac_fit = lavaan::sem(model = OneFacSyntax, 
                  data = cfaData, 
                  fixed.x=FALSE) 
```

request standardized = T to check standardized loadings - item reliability

```{r}
summary(one_fac_fit, standardized = T)
```

+ Only the last three standardized loadings are larger than 0.6
+ The first three indicators are not reliable indicators of the new latent variable eta1

```{r,eval=FALSE}
Model Test User Model:

  Test statistic                               592.661
  Degrees of freedom                                 9
  P-value (Chi-square)                           0.000
```

+ The chi-square statistic is very large and significant for this one-factor model...poor fit

### Plotting

```{r}
semPaths(fixedIndTwoFacRun, what = "std", fade= F)
semPaths(one_fac_fit, what = "std", fade= F)
```

### Comparing Nested Models

+ The one-factor model is nested in the two-factor model. 
+ The fit of the one-factor model is worse than the two-factor model, but is it significantly worse? 
+ Here we use anova() function to perform chi-square difference test
+ Note that the order of the models in anova() doesn't matter 

```{r}
anova(one_fac_fit, fixedIndTwoFacRun)
```

```{r,eval=FALSE}
Chi-Squared Difference Test

                  Df   AIC   BIC    Chisq Chisq diff Df diff Pr(>Chisq)    
fixedIndTwoFacRun  8 14992 15056   2.9575                                  
one_fac_fit        9 15580 15639 592.6611      589.7       1  < 2.2e-16 ***
 ---
 Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
```

+ The model on top is the base model and the model at the bottom is the restricted model.
+ The restricted model always fits worse than the base model. 
+ Chisq diff = 589.7; Df diff = 1; p-value < 0.001
+ Chisq diff is sig: one-factor fits **significantly** worse than the two-factor model and we should endorse two-factor model 

<!------------------------------>
## PART III: Exercises: More Nested Models
<!------------------------------>

Your turn now, Have fun! 

### Exercises: Compare the base model (fixedIndTwoFacRun) to

+ (Model 2) 2-factor CFA model with orthogonal latent variables
+ (Model 3) 2-factor CFA model with a cross-loading from posAffect to satisfied
+ (Model 4) 2-factor CFA model with a correlation between unique factors u1 and u4

### Model 2: Orthogonal Factors

```{r}
OrthFacSyntax <- "
	#Factor Specification	
	posAffect =~ glad + happy + cheerful 
	satisfaction =~ satisfied + content + comfortable
	
	#Orthogonal Factors: no covariance 
	posAffect ~~ 0*satisfaction
"
```

Fit Model 2: 

```{r}
OrthFac_fit <- lavaan::sem(model = OrthFacSyntax, 
                   data = cfaData, 
                   fixed.x = F)
```

Plot Model 2: 

```{r}
semPaths(OrthFac_fit, what = "std", fade= F)
```

chi-square difference test:

```{r}
anova(fixedIndTwoFacRun, OrthFac_fit)
```

+ Chisq diff = 165.52; Df diff = 1; p-value < 0.001
+ Chisq diff is sig: the two-factor model with orthogonal latent variables fits **significantly** worse than the model with correlated latent variables. 
+ We should endorse the base two-factor model with correlated latent variables. 


### Model 3: Cross loading

```{r}
CrossLoadingSyntax <- "
	#Factor Specification	
	# cross loading: satisfied load on both latent variables
	# try to avoid using satisfied as the marker variable
	posAffect =~ glad + satisfied + happy + cheerful 
	satisfaction =~ satisfied + content + comfortable
"
```

Fit Model 3: 

```{r}
CrossLoading_fit <- lavaan::sem(model = CrossLoadingSyntax, 
                                data = cfaData, fixed.x = F)
```

Plot Model 3: 

```{r}
semPaths(CrossLoading_fit, what = "std", fade= F)
```

chi-square difference test:

```{r}
anova(fixedIndTwoFacRun, CrossLoading_fit)
```

+ Chisq diff = 0.052477; Df diff = 1; p-value = 0.8188
+ Chisq diff is not sig: the two-factor model without the cross-loading is **NOT** significantly worse than the model with the cross-loading. 
+ The cross-loading is not necessary. 
+ We should endorse the base two-factor model without the cross-loading. 


### Model 4: Correlated Unique Factors

```{r}
CorrUniSyntax <-"
	#Factor Specification	
	posAffect =~ glad + happy + cheerful 
	satisfaction =~ satisfied + content + comfortable

  # correlated error 
	glad ~~ satisfied
"
```

Fit Model 4: 

```{r}
CorrUni_fit <- lavaan::sem(model= CorrUniSyntax, 
                           data = cfaData, fixed.x = F)
```

Plot Model 4: 

```{r}
semPaths(CorrUni_fit, what = "std", fade= F)
```

chi-square difference test:

```{r}
anova(fixedIndTwoFacRun, CorrUni_fit)
```

+ Chisq diff = 1.3232; Df diff = 1; p-value = 0.25
+ Chisq diff is not sig: the two-factor model without the correlated unique factors is **NOT** significantly worse than the model with the correlated unique factors. 
+ The correlation between u1 and u4 is not necessary. 
+ We should endorse the base two-factor model without the correlated unique factors.


