# Lavaan Lab 10: Model Fit Part II (Fit Indices)


In this lab, we will learn: 

+ how to calculate and interpret global fit indices for SEM models.
+ how to compare non-nested models using AIC and BIC. 

Load up the lavaan library:

```{r}
library(lavaan)
```

<!------------------------------>
## PART I: Fit Indices
<!------------------------------>

Let's read this dataset in: 

```{r}
cfaData<- read.csv("cfaInclassData.csv", header = T)
```

Write out syntax for a two-factor CFA model: 

```{r}
fixedIndTwoFacSyntax <- "
	#Factor Specification	
	posAffect =~ glad + happy + cheerful 
	satisfaction =~ satisfied + content + comfortable
"
```

Fit the model regularly:

```{r}
fixedIndTwoFacRun = lavaan::sem(model = fixedIndTwoFacSyntax, 
                        data = cfaData, 
                        fixed.x=FALSE,
                        estimator = 'MLMV')
```

Request fit indices by adding fit.measures = T in the summary() function:

```{r}
summary(fixedIndTwoFacRun, standardized = T, fit.measures = T)
```

### RMSEA

```{r, eval=FALSE}
Root Mean Square Error of Approximation:

  RMSEA                                          0.000
  90 Percent confidence interval - lower         0.000
  90 Percent confidence interval - upper         0.009
  P-value RMSEA <= 0.05                          1.000
```

reproducing RMSEA: 

+ T = 2.957
+ df = 8
+ N = 1000
+ RMSEA = sqrt(max(T-df,0)/(N-1)/df) = sqrt(max(2.957-8,0)/(1000-1)/8) = 0


### SRMR

```{r, eval=FALSE}
Standardized Root Mean Square Residual:

  SRMR                                           0.007
```

reproducing SRMR: 

```{r}
S = cov(cfaData[,-1])
colnames = colnames(S)
SIGMA = fitted(fixedIndTwoFacRun)$cov[colnames, colnames]
p = ncol(S)

# use cov2cor() function to convert diff to a correlation matrix and standardize the residuals: 

resd = cov2cor(S) - cov2cor(SIGMA)

# keep only the nonduplicated elements:

resd2 = lav_matrix_vech(resd)

sqrt(sum(resd2^2)/(p*(p+1)/2))
```
+ A small average standardized residual...looks good


### Null Model MO

```{r, eval=FALSE}
Model Test Baseline Model:

  Test statistic                              2020.010
  Degrees of freedom                                15
  P-value                                        0.000
```

+ This is the chi_sq for the baseline model used in the CFI/TLI/comparative fit measures.
+ We know what this means now!
+ chisquare of the null model: 2020.010
+ df of the null model: 15

reproducing M0:

```{r}
baselineM0 <- "
	glad ~~ glad
	happy ~~ happy
	cheerful ~~ cheerful
	satisfied ~~ satisfied
	content ~~ content
	comfortable ~~ comfortable
"
base_fit <- lavaan::sem(baselineM0, data = cfaData, fixed.x = FALSE)
base_fit
```


### CFI/TLI

```{r, eval=FALSE}
User Model versus Baseline Model:

  Comparative Fit Index (CFI)                    1.000
  Tucker-Lewis Index (TLI)                       1.005
```


+ 100% improvement over the null(baseline) model ... great fit
+ Here TLI is larger than 1 because this is a rare situation with chisquare=2.957<df=8
+ numerator of TLI = (2020.010/15-2.957/8) = 134.2977
+ denominator of TLI = (2020.010/15-1) = 133.6673
+ TLI = 134.2977 / 133.6673 = 1.005
+ A TLI that is larger than 1 is no different from TLI = 1

great overall fit. 

### Loglikelihood

```{r, eval=FALSE}
Loglikelihood and Information Criteria:

  Loglikelihood user model (H0)              -7483.272
  Loglikelihood unrestricted model (H1)      -7481.793
```

+ H0 is the loglikelihood of your model (user model ... lavaan is kind to clarify this)
+ H1 is the saturated model loglikelihood.

### AIC/BIC

```{r, eval=FALSE}
  Akaike (AIC)                               14992.544
  Bayesian (BIC)                             15056.345
  Sample-size adjusted Bayesian (BIC)        15015.056
```

+ Penalized -2 LogL
+ If these are lower than some other model -> prefer this model.
+ If these are higher than some other model -> prefer the other model.

reproducing AIC/BIC:

```{r}
logLik = -7483.272
q = 13 # (4 loadings + 6 unique factor variances + 3 factor var/covs)
N = 1000

(AIC = -2*logLik + 2*q)
(BIC = -2*logLik + log(N)*q)
```

<!------------------------------>
## PART II: Exercise
<!------------------------------>

For this portion, we will run the CFA analyses on a new simulated dataset based on Todd Little's positive affect example. 

Read in the new dataset:

```{r}
affectData_new <- read.csv("ChiStatSimDat.csv", header = T)
```

Examine the dataset:

```{r}
head(affectData_new)
```

Examine the covariance matrix:

```{r}
cov(affectData_new)
```
all positive! (Remember that indicators need to be all positively correlated for CFA models?)

### PART I: Plot the distributions of all indicators

```{r}
library(PerformanceAnalytics)
chart.Correlation(affectData_new)
```

all indicators look roughly normal

### PART II: Write out the model syntax for two-factor model

```{r}
twofa.model <- "
	#Factor Specification	
	posAffect =~ glad + happy + cheerful 
	satisfaction =~ satisfied + content + comfortable  
"
```

### PART III: Fit the two-factor model

```{r}
new_fit = lavaan::sem(twofa.model, data = affectData_new, fixed.x=FALSE)
summary(new_fit, standardized = T, fit.measures = T)
```

### PART IV: Interpret the chisquare statistic and fit indices

```{r,eval=FALSE}
Model Test User Model:
                                                      
  Test statistic                               113.638
  Degrees of freedom                                 8
  P-value (Chi-square)                           0.000
```

+ 113.638 is much larger than df=8 and the p-value is 0.000<0.05,
+ This chisquare is too large and the model is a poor fit. 

```{r,eval=FALSE}
Root Mean Square Error of Approximation:

  RMSEA                                          0.257
  90 Percent confidence interval - lower         0.216
  90 Percent confidence interval - upper         0.300
  P-value RMSEA <= 0.05                          0.000

Standardized Root Mean Square Residual:

  SRMR                                           0.070

User Model versus Baseline Model:

  Comparative Fit Index (CFI)                    0.908
  Tucker-Lewis Index (TLI)                       0.828
```

+ RMSEA = 0.257 >> 0.1. The lower bound of the confidence interval is also larger than 0.1. This indicates a poor fit. 
+ P-value RMSEA <= 0.05: sig - close fit null hypothesis rejected.
+ SRMR = 0.07 < 0.08, meaning that the average standardized residual between S and Sigma is no larger than 0.08, but SRMR is known to be lenient (i.e., low SRMR =/= good models, but high SRMR = bad models). 
+ 90.8% improvement over the null model ... marginal fit


