# Lavaan Lab 11: Model Local Fitting and Model Modifications


In this lab, we will learn: 

+ how to examine SEM local fit using residuals 
+ how to modify SEM models for improved fit using modification indices

Load up the lavaan library:

```{r}
library(lavaan)
```

<!------------------------------>
## PART I: Local Fit with Residuals
<!------------------------------>

Let's read in the new dataset ChiStatSimDat.csv: 

```{r}
cfaData<- read.csv("ChiStatSimDat.csv", header = T)
```

Write out syntax for a two-factor CFA model: 

```{r}
fixedIndTwoFacSyntax <- "
	#Factor Specification	
	posAffect =~ glad + happy + cheerful 
	satisfaction =~ satisfied + content + comfortable
"
```

Fit the two-factor model:

```{r}
fixedIndTwoFacRun = lavaan::sem(model = fixedIndTwoFacSyntax, 
                        data = cfaData, 
                        fixed.x=FALSE)
# , estimator = 'MLMV'
summary(fixedIndTwoFacRun, standardized = T, fit.measures = T)

```

### Unstandardized residuals

```{r}
resid(fixedIndTwoFacRun)$cov
```

What does this mean? What is the metric?

### Standardized residuals

```{r}
resid(fixedIndTwoFacRun, type = "standardized")$cov
```

### Normalized residuals

```{r}
resid(fixedIndTwoFacRun, type = "normalized")$cov
```

+ Different residuals, same story
+ The covariance residual between cheerful and comfortable is the largest and positive
+ The model under-predicts this covariance
+ Fix!


<!------------------------------>
## PART II: Modification Indices
<!------------------------------>

```{r, eval=FALSE}
modindices(fixedIndTwoFacRun)
```

Filter output and only show rows with a modification index value equal or higher than 1:

```{r}
modindices(fixedIndTwoFacRun, minimum.value = 10)
```

Sort the output using the values of the modification index values. Higher values appear first: 
```{r}
modindices(fixedIndTwoFacRun, minimum.value = 10, sort = TRUE)
```

+ op ~~ : a correlation between two unique factors 
+ op =~ : cross-loading
+ This indicates that the parameters lavaan detects for you to free up are all residual covariances.

### Modified Model 1:

```{r}
mod1 <- "
	posAffect =~ glad + happy + cheerful 
	satisfaction =~ satisfied + content + comfortable  
	
	#residal covariance
	cheerful ~~ comfortable
"
```

```{r}
mod1_fit <- lavaan::sem(mod1, data = cfaData, 
                        std.lv = TRUE, fixed.x=FALSE)

summary(mod1_fit, fit.measures = T, standardized = T)
```

Model comparison: 

```{r}
anova(mod1_fit, fixedIndTwoFacRun)
```

Keep modifying mod1:

```{r}
modindices(mod1_fit, minimum.value = 10, sort = TRUE)
```


### Modified Model 2_1:

```{r}
mod2_1 <- "
  # cross loading
	posAffect =~ glad + happy + cheerful 
	satisfaction =~ cheerful + satisfied + content + comfortable  # cheerful also loads on satisfaction
	
	#residal covariance
	cheerful ~~ comfortable
"
```

```{r}
mod2_1_fit <- lavaan::sem(mod2_1, data = cfaData, 
                          std.lv = TRUE, fixed.x=FALSE)

summary(mod2_1_fit, fit.measures = T, standardized = T)
```

Model comparison: 

```{r}
anova(mod2_1_fit, mod1_fit)
```


### Modified Model 2_2:

```{r}
mod2_2 <- "
	posAffect =~ glad + happy + cheerful 
	satisfaction =~ satisfied + content + comfortable 
	
	#residal covariance
	cheerful ~~ comfortable
	glad ~~ happy
"
```

```{r}
mod2_2_fit <- lavaan::sem(mod2_2, data = cfaData, 
                          std.lv = TRUE, fixed.x=FALSE)

summary(mod2_2_fit, fit.measures = T, standardized = T)
```

Model comparison: 

```{r}
anova(mod2_2_fit, mod1_fit)
```

Keep modifying 2_2:

```{r}
modindices(mod2_2_fit, minimum.value = 10, sort = TRUE)
```



### Modified Model 3:

```{r}
mod3 <- "
	posAffect =~ glad + happy + cheerful 
	satisfaction =~ satisfied + content + comfortable 
	
	#residal covariance
	cheerful ~~ comfortable
	glad ~~ happy
	content ~~ happy
	glad ~~ content
"
```

```{r}
mod3_fit <- lavaan::sem(mod3, data = cfaData, 
                        std.lv = TRUE, fixed.x = FALSE)

summary(mod3_fit, fit.measures = T, standardized = T)
```

Model comparison: 

```{r}
anova(mod3_fit, mod2_2_fit)
```

Keep modifying mod3:

```{r}
modindices(mod3_fit, minimum.value = 1, sort = TRUE)
```

No suggestions could decrease the model chisquare by more than 10. 