# Lavaan Lab 13: SEM for Nonnormal and Categorical Data

Load up the lavaan library:

```{r}
library(lavaan)
```

<!------------------------------>
## PART I: Nonnormality Diagnosis
<!------------------------------>

Let's first load the simulated non-normal data and look at the normality/nonnormality of the items:

```{r}
nnorm_dat <- read.csv("nonnormal.csv", header = T)
head(nnorm_dat)
```

```{r}
par(mfrow = c(2, 2)) #opens graph window with 2 rows 2 columns
hist(nnorm_dat$odd5)
hist(nnorm_dat$odd6)
hist(nnorm_dat$odd7)
hist(nnorm_dat$odd8)
```

Use describe() function from the psych package to get univariate descriptives:

```{r,message=FALSE}
#install.packages("psych")
library(psych)
describe(nnorm_dat)
```

Use mardia() from the psych package to test multivariate normality:

```{r}
par(mfrow = c(1, 1)) #opens graph window
mardia(nnorm_dat)
```

In any case, these data are clearly far from normal, so ...

<!------------------------------>
## PART II: Robust corrections
<!------------------------------>

Write out syntax for a one-factor CFA model: 

```{r}
cfaSyn <- "
	odd =~ odd1 + odd2 + odd3 + odd4 + odd5 + odd6 + odd7 + odd8
	"
```

Fit the one-factor model:

```{r}
mlrFit <- lavaan::sem(cfaSyn, 
              data = nnorm_dat, 
              fixed.x = FALSE,
              estimator = "mlr")

summary(mlrFit, fit.measure = T)
```


<!------------------------------>
## PART III: Categorical Data Analysis in Lavaan
<!------------------------------>

Let's load the simulated data in which ODD items are ordinal:

```{r}
odd <- read.csv("oddData.csv", header = T)
head(odd)
```

Write out syntax for a one-factor CFA model: 

```{r}
oddOneFac = '
	#Specify Overall Odd Factor
	odd  =~ odd1 + odd2 + odd3 + odd4 + odd5 + odd6 + odd7 + odd8
	'
```

+ Fit the one-factor model:
+ label ordinal variables using ordered argument: 
+ ordered = c( #NAMES OF ORDINAL INDICATORS#)

```{r}
oneFacFit <- lavaan::sem(oddOneFac, 
                 data = odd,
                 ordered=c('odd1','odd2','odd3','odd4','odd5','odd6','odd7','odd8'), 
                 fixed.x = FALSE,
                 parameterization = 'theta')  # or delta 
#declare these as ordered variable

summary(oneFacFit, fit.measures = T)
```


<!------------------------------>
## PART IV: What if you have it all? 
<!------------------------------>

+ Unfortunately you cannot use missing = 'fiml' for categorical data:

```{r}
FitMessy <- lavaan::sem(oddOneFac, 
                data = odd,
                ordered=c('odd1','odd2','odd3','odd4','odd5','odd6','odd7','odd8'), 
                fixed.x = FALSE,
                estimator = "DWLS",
                #missing = 'fiml'
                ) 

FitMessy
#summary(FitMessy, fit.measures = T)
```

+ But you cannot use missing = 'fiml' together with MLR for nonnormal data:

```{r}
FitMessy <- lavaan::sem(oddOneFac, 
                data = nnorm_dat,
                #ordered=c('odd1','odd2','odd3','odd4','odd5','odd6','odd7','odd8'), 
                fixed.x = FALSE,
                estimator = "mlr",
                missing = 'fiml') 

FitMessy
#summary(FitMessy, fit.measures = T)
```
