# Lavaan Lab 16: Latent Growth Models

In this lab, we will:

+ run and interpret a series of growth models (no growth, linear, quadratic, latent basis, spline growth);
+ compare nested models and identify the best possible shape for characterizing the growth patterns; 
+ add predictors for the growth factors;
+ run growth models on latent variables. 

Load up the lavaan library:

```{r, message=FALSE}
library(lavaan)
```

We will also need ggplot2, semPlot, and semTools. Install them if you haven't: 

```{r,message=FALSE}
#install.packages("ggplot2")
#install.packages("semPlot")
#install.packages("semTools")
library(ggplot2)
library(semPlot)
library(semTools)
```

+ For this lab, we will work with a simulated dataset # based on an example from McCoach & Kaniskan (2010). 
+ The main DV is Oral Reading Fluency (ORF) and is measured over 4 time points (Fall and Spring, 2 consecutive years) 
+ N = 277 Elementary students.
+ Let's read in the dataset:

```{r}
orf <- read.csv("readingSimData.csv", header = T)
```

Take a look at the dataset:

```{r}
head(orf)
```

sample size:
```{r}
n <- nrow(orf)
n #277, just like the McCoach paper.
```

sample means and cov matrix
```{r}
orfNames <- paste0("orf", 1:4)
(samMeans <- round(apply(orf[,orfNames], 2, mean), 3))
(samCov <- round(cov(orf[,orfNames])*((n-1)/n), 3))
```

<!------------------------------>
## PART I: Spaghetti Plot
<!------------------------------>

For more details, check out https://www.r-bloggers.com/my-commonly-done-ggplot2-graphs/

First, let's use reshape() to convert wide format to long format for plotting: 

```{r}
growthDataLong <- reshape(orf, varying = paste0("orf", 1:4), sep = "", direction = "long")
head(growthDataLong)
```

Plot trajectory of individual with id=1

```{r}
tspag_id1 = ggplot(growthDataLong[growthDataLong$id==1, ], aes(x=time, y=orf)) + 
  geom_line() + 
  xlab("Observation Time Point") +
  ylab("Y") + 
  ylim(-40, 300) + 
  ggtitle("Spaghetti plot") + 
  aes(colour = factor(id))
tspag_id1
```

plot trajectory of everyone 

```{r}
tspag = ggplot(growthDataLong, aes(x=time, y=orf)) + 
  geom_line(show.legend = FALSE) + 
  xlab("Observation Time Point") +
  ylab("Y") + 
  ylim(-40, 300) + 
  ggtitle("Spaghetti plot") + 
  aes(colour = factor(id))
tspag
```


<!------------------------------>
## PART II: Growth Models
<!------------------------------>

### 1. No growth model

+ Let's start by examining the hypothesis of no growth (intercept only)
+ Intercept loads on all variables with fixed loadings of 1.0
+ Use a*VAR1 to fix the coefficient of VAR1 at a:

```{r}
noGrowthSyn <- "
	#Specify Latent Intercept
	I =~ 1*orf1 + 1*orf2 + 1*orf3 + 1*orf4
"
```

```{r}
noGrowthFit <- growth(noGrowthSyn, data = orf, fixed.x = FALSE)
summary(noGrowthFit, fit.measures = T)
```

```{r}
semPaths(noGrowthFit, what='est', fade= F)
```

### 2. Linear growth model

+ Intercept loads on all variables with fixed loadings of 1.0
+ Slope loads on all variables with fixed loadings of t = 0, 1, 2, ..., t-1
+ t must start from 0

```{r}
linearGrowthSyn <- "
	#Specify Latent Intercept and Slope
	I =~ 1*orf1 + 1*orf2 + 1*orf3 + 1*orf4
	S =~ 0*orf1 + 1*orf2 + 2*orf3 + 3*orf4
"
```

```{r}
linearGrowthFit <- growth(linearGrowthSyn, data = orf, fixed.x = FALSE)
summary(linearGrowthFit, fit.measures = T)
```

```{r}
semPaths(linearGrowthFit, what='est', fade= F)
```

### 3. Quadratic growth model

+ Intercept loads on all variables with fixed loadings of 1.0
+ Slope loads on all variables with fixed loadings of t = 0, 1, 2, ..., t-1
+ Quadratic loads on all variables with fixed loadings of t^2 = 0, 1, 4, ..., (t-1)^2
+ Quadratic has no variance and covariances

```{r}
quadGrowthSyn <- "
	#int and slope factors
	I =~ 1*orf1 + 1*orf2 + 1*orf3 + 1*orf4
	S =~ 0*orf1 + 1*orf2 + 2*orf3 + 3*orf4
	#quadratic factor = slope^2
	quadS =~ 0*orf1 + 1*orf2 + 4*orf3 + 9*orf4
"
```

```{r}
quadGrowthFit <- growth(quadGrowthSyn, data = orf, fixed.x = FALSE)
```

If you get the following warning messages:

1: In lav_object_post_check(object) :
  lavaan WARNING: some estimated ov variances are negative

+ Use var1~~0*var2 to fix the (co)variances at 0

```{r}
quadGrowthSyn_noQuad <- "
	#int and slope factors
	I =~ 1*orf1 + 1*orf2 + 1*orf3 + 1*orf4
	S =~ 0*orf1 + 1*orf2 + 2*orf3 + 3*orf4
	#quadratic factor = slope^2
	quadS =~ 0*orf1 + 1*orf2 + 4*orf3 + 9*orf4
	
	quadS ~~ 0*quadS #restrict quadratic variance to 0
	quadS ~~ 0*I #restrict quadratic covariance with I to 0
	quadS ~~ 0*S #restrict quadratic covariance with S to 0
"
```

```{r}
quadGrowthNoquadFit <- growth(quadGrowthSyn_noQuad, 
                              data = orf, fixed.x = FALSE)
summary(quadGrowthNoquadFit, fit.measures = T)
```

```{r}
semPaths(quadGrowthNoquadFit, what='est', fade= F)
```

### 4. Latent basis growth model (extension of linear growth model)

+ Intercept loads on all variables with fixed loadings of 1.0
+ Slope loads on all variables with free loadings between 0 and t-1

```{r}
latentBasisSyn <- "
	#Int and slope specification
	I =~ 1*orf1 + 1*orf2 + 1*orf3 + 1*orf4
	
	#orf2 and orf3 are free in the latent basis specification
	S =~ 0*orf1 + alpha1*orf2 + alpha2*orf3 + 3*orf4
"
```

```{r}
latentBasisFit <- growth(latentBasisSyn, 
                         data = orf, 
                         fixed.x = FALSE)
summary(latentBasisFit, fit.measures = T)
```

```{r}
semPaths(latentBasisFit, what='est', fade= F)
```

RMSEA failed us...
one approach is to use model modification indices: 

```{r,warning=FALSE}
modindices(latentBasisFit,sort. = T)
```

Another approach is to use Spline Growth Model.

### 5. Spline Growth Model

```{r}
splineGrowthSyn <- "
	#Specify Latent Intercept and Slope
	I =~ 1*orf1 + 1*orf2 + 1*orf3 + 1*orf4
	S =~ 0*orf1 + 1*orf2 + 2*orf3 + 3*orf4

	#summer is the spline variable
	summer =~ 0*orf1 + 0*orf2 + 1*orf3 + 1*orf4

	#Summer gets a mean but no variance
	summer ~ 1
	summer ~~ 0*summer

	#Summer is uncorrelated with I and S
	summer ~~ 0*I 
	summer ~~ 0*S 
"
```

```{r}
splineGrowthFit <- growth(splineGrowthSyn, data = orf, fixed.x = FALSE)
summary(splineGrowthFit, fit.measures = T)
```

```{r}
semPaths(splineGrowthFit, what='est', fade= F)
```

### Model Comparison

```{r}
lavTestLRT(noGrowthFit, linearGrowthFit)
```

+ Linear Growth Model fits significantly better than No Growth Model

```{r}
lavTestLRT(linearGrowthFit, quadGrowthNoquadFit)
```

+ Linear Growth Model fits almost the same as the Quadratic Growth Model (keep linear model due to parsimony principle)

```{r}
lavTestLRT(linearGrowthFit, latentBasisFit)
```

+ Latent Basis Model fits significantly better than Linear Growth Model

```{r}
lavTestLRT(linearGrowthFit, splineGrowthFit)
```

+ Spline Growth Model also fits significantly better than Linear Growth Model

### 6. Final Model: Spline Growth Model with a binary treatment predictor

```{r}
splineGrowthTreatmentPredictorSyn <- "
	#Specify Latent Intercept and Slope
	I =~ 1*orf1 + 1*orf2 + 1*orf3 + 1*orf4
	S =~ 0*orf1 + 1*orf2 + 2*orf3 + 3*orf4

	#summer is the spline variable
	summer =~ 0*orf1 + 0*orf2 + 1*orf3 + 1*orf4

	#Summer gets a mean but no variance
	summer ~~ 0*summer

	#Summer is uncorrelated with I and S
	summer ~~ 0*I 
	summer ~~ 0*S 

	#Intercept, Slope, and Summer regressed on (predicted by) treatment
	I ~ treatmentDummy
	S ~ treatmentDummy
	summer ~ treatmentDummy
"
```

+ When including external predictors, we need to turn on fixed.x = T...
+ otherwise you'll get a warning message and misleading model fit:

```{r}
# do not do this:
splineGrowthTreatPredictorFit <- growth(splineGrowthTreatmentPredictorSyn, 
                                       data = orf, 
                                       fixed.x = F) # <- Here
```

Instead, turn on fixed.x = T:

```{r}
splineGrowthTreatPredictorFit <- growth(splineGrowthTreatmentPredictorSyn, 
                                       data = orf, 
                                       fixed.x = T) # <- Here
summary(splineGrowthTreatPredictorFit, fit.measures = T)
```

```{r}
semPaths(splineGrowthTreatPredictorFit, what='est', fade= F)
```


<!------------------------------>
## PART III: LGM on Latent Variables
<!------------------------------>

### Example

Please go over the checklist:

1. Make sure the latent variables satisfy longitudinal measurement invariance at the level of scalar invariance or above 
2. Use the loadings over time (i.e., metric invariance)
3. No need to correlate the latent factors
4. Add intercepts for all indicators EXCEPT for marker indicators
5. Add correlated residuals for repeated measures of the same indicators
6. Use std.lv = TRUE as the scaling method in growth()

For this example, we will use the dataset exLong from package semTools: 

```{r}
data(exLong)
head(exLong)
?exLong
```

The syntax for linear growth model with latent variables:

```{r}
exLinearGrowthsyn <- "
  
  # Use the loadings over time (i.e., metric invariance)
  f_t1 =~ lamb1*y1t1 + lamb2*y2t1 + lamb3*y3t1
  f_t2 =~ lamb1*y1t2 + lamb2*y2t2 + lamb3*y3t2
  f_t3 =~ lamb1*y1t3 + lamb2*y2t3 + lamb3*y3t3

  #Int and slope specification
	I =~ 1*f_t1 + 1*f_t2 + 1*f_t3
	S =~ 0*f_t1 + 1*f_t2 + 2*f_t3
	
	# Add intercepts for all indicators EXCEPT for marker indicators
  y2t1 ~ 1
  y3t1 ~ 1
  y2t2 ~ 1
  y3t2 ~ 1
	y2t3 ~ 1
  y3t3 ~ 1

	# Add correlated residuals for repeated measures of the same indicators
	y1t1 ~~ y1t2
	y1t1 ~~ y1t3
	y1t2 ~~ y1t3
	y2t1 ~~ y2t2
	y2t1 ~~ y2t3
	y2t2 ~~ y2t3
	y3t1 ~~ y3t2
	y3t1 ~~ y3t3
	y3t2 ~~ y3t3
"
```

Use std.lv = TRUE as the scaling method in growth():

```{r}
exLinearGrowthFit <- growth(exLinearGrowthsyn, 
                         data = exLong, 
                         fixed.x = FALSE,
                         std.lv = TRUE)
summary(exLinearGrowthFit, fit.measures = T)
```

```{r}
semPaths(exLinearGrowthFit, what='est', fade= F)
```

### Exercise

Q: Could you fit the latent basis model to the same dataset and compare the fit of the two models?

```{r}

```

