# Lavaan Lab 17: Second-order and Bifactor Models

In this lab, we will evaluate the dimensionality of ISMI-29 by fitting and comparing the following four models:

1. Unidimensional model (one-factor CFA)
2. Correlated factors model (multi-factor CFA)
3. Second-order factor model
4. Bifactor model 

Load up the lavaan and semPlot libraries:

```{r, message=FALSE}
library(lavaan)
library(semPlot)
```

+ In this lab, we will work with the ISMI-29 data that are collected using Internalized Stigma of Mental Illness Scale 
+ 758 participants and 29 items
+ Let's read in the dataset:

```{r}
ISMI29 = read.csv('ISMI-29 n758 (Hammer 16).csv', header = F)
```

Take a look at the dataset:

```{r}
head(ISMI29)
```

sample size:
```{r}
n <- nrow(ISMI29)
n #758
```

Factor structure: 

+ Item1-6: Alienation  “Having a mental illness has spoiled my life.”
+ Item7-13: Stereotype Endorsement  “Mentally ill people tend to be violent”
+ Item14-18: Discrimination Experience   “People discriminate against me because I have a mental illness”
+ Item19-24: Social Withdrawal   “I don’t talk about myself as much because I don’t want to burden others with my mental illness”
+ Item25-29: \*Stigma Resistance  (*reverse-coded) “I can have a good, fulfilling life, despite my mental illness”    

<!------------------------------>
## PART I: Unidimensional model
<!------------------------------>

Write out syntax for a one-factor CFA model: 

```{r}
uni.model = '
ISMI =~ V1+V2+V3+V4+V5+V6+V7+V8+V9+V10+V11+V12+V13+V14+V15+V16+V17+V18+V19+
        V20+V21+V22+V23+V24+V25+V26+V27+V28+V29
'
```

Fit the model:

+ It is recommended to fix the variances of all first- and second-order factors to be 1 (lavaan: std.lv = TRUE) and request standardized solutions;

```{r}
uni.model.fit = lavaan::sem(uni.model, 
                    data=ISMI29, 
                    ordered = colnames(ISMI29), 
                    std.lv = TRUE, 
                    fixed.x = F)
summary(uni.model.fit, standardized = TRUE, fit.measures = TRUE)
```

Plot the path diagram:

```{r}
semPaths(uni.model.fit, what = 'std', fade = F)
```

<!------------------------------>
## PART II: Correlated factors model
<!------------------------------>

Write out syntax for a five-factor CFA model: 

```{r}
cor.fac.model = '
Alienation =~ V1+V2+V3+V4+V5+V6
Stereotype =~ V7+V8+V9+V10+V11+V12+V13
Discrimination =~ V14+V15+V16+V17+V18
Withdrawal =~ V19+V20+V21+V22+V23+V24
Stigma =~ V25+V26+V27+V28+V29
'
```

```{r}
cor.fac.model.fit = lavaan::sem(cor.fac.model, 
                        data=ISMI29, 
                        ordered = colnames(ISMI29), 
                        std.lv = TRUE, 
                        fixed.x = F)

summary(cor.fac.model.fit, standardized = TRUE, fit.measures = TRUE)
```

```{r}
semPaths(cor.fac.model.fit, what = 'std', fade = F)
```


<!------------------------------>
## PART III: Second-order factor Model
<!------------------------------>

Write out syntax for a five-factor second-order CFA model: 

```{r}
secondfac.model = '
Alienation =~ V1+V2+V3+V4+V5+V6
Stereotype =~ V7+V8+V9+V10+V11+V12+V13
Discrimination =~ V14+V15+V16+V17+V18
Withdrawal =~ V19+V20+V21+V22+V23+V24
Stigma =~ V25+V26+V27+V28+V29

# Second-order factor ISMI
ISMI =~ Alienation + Stereotype + Discrimination + Withdrawal + Stigma
'
```

```{r}
secondfac.model.fit = lavaan::sem(secondfac.model, 
                          data=ISMI29, 
                          ordered = colnames(ISMI29), 
                          std.lv = TRUE, 
                          fixed.x = F)
summary(secondfac.model.fit, standardized = TRUE, fit.measures = TRUE)
```

```{r}
semPaths(secondfac.model.fit, what = 'std', fade = F)
```

<!------------------------------>
## PART IV: Bifactor Model
<!------------------------------>

```{r}
bifac.model = '
# specific factors
Alienation =~ V1+V2+V3+V4+V5+V6
Stereotype =~ V7+V8+V9+V10+V11+V12+V13
Discrimination =~ V14+V15+V16+V17+V18
Withdrawal =~ V19+V20+V21+V22+V23+V24
Stigma =~ V25+V26+V27+V28+V29

# general factor GEN
GEN =~ V1+V2+V3+V4+V5+V6+V7+V8+V9+V10+V11+V12+V13+V14+V15+V16+V17+V18+V19+
      V20+V21+V22+V23+V24+V25+V26+V27+V28+V29
'
```

When using sem() to fit a bifactor model, make sure to turn on 

+ orthogonal = TRUE to ensure that all specific factors and general factors are uncorrelated
+ otherwise, you'll get an error/warning saying that the model is not identified.

```{r}
bifac.model.fit = lavaan::sem(bifac.model, 
                      data=ISMI29, 
                      ordered = colnames(ISMI29), 
                      std.lv = TRUE, 
                      fixed.x = F,
                      orthogonal = TRUE)
```

```{r}
summary(bifac.model.fit, standardized = TRUE, fit.measures = TRUE)
```

```{r}
semPaths(bifac.model.fit, what = 'std', fade = F)
```


<!------------------------------>
## PART V: Model Comparison 
<!------------------------------>

```{r}
UniFactor = fitMeasures(uni.model.fit, fit.measures = c("chisq.scaled", "df.scaled", "pvalue.scaled", "rmsea.scaled", "rmsea.ci.lower.scaled", "rmsea.ci.upper.scaled", "cfi.scaled", "tli.scaled", "srmr_bentler"))
FiveFactor = fitMeasures(cor.fac.model.fit, fit.measures = c("chisq.scaled", "df.scaled", "pvalue.scaled", "rmsea.scaled", "rmsea.ci.lower.scaled", "rmsea.ci.upper.scaled", "cfi.scaled", "tli.scaled", "srmr_bentler"))
SecondOrder = fitMeasures(secondfac.model.fit, fit.measures = c("chisq.scaled", "df.scaled", "pvalue.scaled", "rmsea.scaled", "rmsea.ci.lower.scaled", "rmsea.ci.upper.scaled", "cfi.scaled", "tli.scaled", "srmr_bentler"))
Bifactor = fitMeasures(bifac.model.fit, fit.measures = c("chisq.scaled", "df.scaled", "pvalue.scaled", "rmsea.scaled", "rmsea.ci.lower.scaled", "rmsea.ci.upper.scaled", "cfi.scaled", "tli.scaled", "srmr_bentler"))
round(cbind(UniFactor, FiveFactor, SecondOrder, Bifactor), 3)
```

Bifactor model wins!

Bifactor indices:

```{r}
library(BifactorIndicesCalculator)
bifactorIndices(bifac.model.fit)
```


<!------------------------------>
## Exercise: Mental Ability Scale
<!------------------------------>

Let's bring our Holzinger and Swineford Dataset back: 

```{r}
head(HolzingerSwineford1939)
?HolzingerSwineford1939
```

This dataset has 301 cases with 9 mental ability items.

**Assignment: Could you use the four models above to examine the dimensionality of this ODD Subscale?**

Here is a factor structure that you may need: 

```{r}
cor.fac.HS.model = '
visual  =~ x1 + x2 + x3
textual =~ x4 + x5 + x6
speed   =~ x7 + x8 + x9
'
```

Good luck!

```{r}
bifac.model = '
# specific factors
visual  =~ x2 + x3 # remove x1 because of heywood case
textual =~ x4 + x5 + x6
speed   =~ x7 + x8 + x9

# general factor GEN <- 
G =~ x1 + x2 + x3 + x4 + x5 + x6 + x7 + x8 + x9
'
bifac.model.fit = lavaan::sem(bifac.model, 
                      data=HolzingerSwineford1939, 
                      #ordered = colnames(odd), 
                      std.lv = TRUE, 
                      fixed.x = F,
                      orthogonal = TRUE,
                      estimator = 'MLR')
lavaan:::summary(bifac.model.fit, standardized = TRUE, fit.measures = TRUE)
```

