# Lavaan Lab 18: CFA of MTMM Matrix

In this lab, we will:
    
+ run CFA on MTMM Matrix to investigate convergent and discrimative validity 

Load up the lavaan and semPlot libraries:

```{r, message=FALSE}
library(lavaan)
library(semPlot)
```

+ Let's read in a simulated MTMM matrix:

```{r}
load("MTMM.RData")
```

Take a look at the matrix:

```{r}
dim(MTMM)
head(MTMM)
```
This is a covariance matrix. You could also convert it to a correlation matrix: 

```{r}
cov2cor(MTMM)
```


<!------------------------------>
## PART I: Correlated methods specification 
<!------------------------------>

This model specifies both traits and methods factors:

```{r}
MTMM.model.spec1.wrong <- '
# trait factors

paranoid =~ PARI +  PARC + PARO   
schizotypal =~ SZTI + SZTC + SZTO
schizoid =~ SZDI + SZDC + SZDO

# method factors

inventory =~ SZTI + PARI + SZDI
clininter =~ PARC + SZTC + SZDC
obsrating =~ PARO + SZTO + SZDO
'
```

Fit the model:

+ Since MTMM is a covariance matrix, we supply the sample size 500;

```{r}
fit1 <- lavaan::sem(MTMM.model.spec1.wrong, 
            sample.cov=MTMM, sample.nobs=500,
            fixed.x = F)
```

You might get the following warning message:

```{r, eval=FALSE}
Warning message:
  In lavaan::lavaan(model = MTMM.model.spec1.wrong, sample.cov = MTMM,  :
        lavaan WARNING:
        the optimizer (NLMINB) claimed the model converged, but not all
        elements of the gradient are (near) zero; the optimizer may not
        have found a local solution use check.gradient = FALSE to skip
        this check.                        
```

+ The problem is by default lavaan correlates all traits and methods factors; 
+ To get the model to fit, we need to manually uncorrelate traits and methods factors;

```{r}
MTMM.model.spec1 <- '
# trait factors

paranoid =~ PARI +  PARC + PARO   
schizotypal =~ SZTI + SZTC + SZTO
schizoid =~ SZDI + SZDC + SZDO

# method factors

inventory =~ SZTI + PARI + SZDI
clininter =~ PARC + SZTC + SZDC
obsrating =~ PARO + SZTO + SZDO

# uncorrelated trait and method

paranoid ~~ 0*inventory
paranoid ~~ 0*clininter
paranoid ~~ 0*obsrating
schizotypal ~~ 0*inventory
schizotypal ~~ 0*clininter
schizotypal ~~ 0*obsrating
schizoid ~~ 0*inventory
schizoid ~~ 0*clininter
schizoid ~~ 0*obsrating
'
```

Model fit:

```{r}
fit2 <- lavaan::sem(MTMM.model.spec1, 
            sample.cov=MTMM, sample.nobs=500,
            fixed.x = F)

summary(fit2, standardized = T, fit.measures = T)
```

```{r, eval=FALSE}
Heywood case
Warning messages:
1: In lav_object_post_check(object) :
  lavaan WARNING: some estimated ov variances are negative
```

Plot the path diagram:

```{r}
semPaths(fit2, what='std', 
         nCharNodes = 0,
         nCharEdges = 0, # don't limit variable name lengths
         curvePivot = TRUE, 
         curve = 1.1, # pull covariances' curves out a little
         fade=FALSE)
```


<!------------------------------>
## PART II: Correlated uniqueness specification
<!------------------------------>

In this specification: 

+ There is no method factor;
+ Instead, the unique factors are correlated within method blocks; 

```{r}
MTMM.model.spec2 <- '
# trait factors

paranoid =~ PARI + PARC + PARO   
schizotypal =~ SZTI + SZTC + SZTO
schizoid =~ SZDI + SZDC + SZDO

# no method factors

# correlated residual covariances

# Method 1 Block
PARI ~~ SZTI + SZDI
SZTI ~~ SZDI

# Method 2 Block
PARC ~~ SZTC + SZDC
SZTC ~~ SZDC

# Method 3 Block
PARO ~~ SZTO + SZDO
SZTO ~~ SZDO
'
```

Model fit:

```{r}
fit3 <- lavaan::sem(MTMM.model.spec2, 
            sample.cov=MTMM, sample.nobs=500,
            fixed.x = F, std.lv = T)

#results with standardized parameter estimates
summary(fit3, standardized=TRUE, fit.measures=TRUE)
```

```{r}
semPaths(fit3, what='std', 
         nCharNodes = 0,
         nCharEdges = 0, # don't limit variable name lengths
         curvePivot = TRUE, 
         curve = 1.1, # pull covariances' curves out a little
         fade=FALSE)
```


