# Lavaan Lab 19: Multilevel SEM

In this lab, we will:
    
+ build a multilevel CFA model 
+ add covariates at both the between and the within level

Load up the lavaan library:

```{r, message=FALSE}
library(lavaan)
```

Let's read in a Mplus example dataset from an online location

```{r}
Data <- read.table("http://statmodel.com/usersguide/chap9/ex9.6.dat")
names(Data) <- c("y1", "y2", "y3", "y4", "x1", "x2", "w", "clus")
```

Take a look at the matrix:

```{r}
head(Data)
dim(Data)
length(unique(Data$clus))
```

+ there are 1000 individual observations in 110 clusters
+ cluster sizes: 5, 10, 15
+ 4 measures at the within level y1, y2, y3, y4
+ 2 covariates at the within level: x1, x2
+ 1 covariate at the between level: w

<!------------------------------>
## PART I: Multilevel CFA 1: within-only construct
<!------------------------------>

This model specifies the latent variable only at the within level:

```{r}
model1 <- '
  level: 1
    fw =~ y1 + y2 + y3 + y4

  level: 2
    y1 ~~ y1 + y2 + y3 + y4
    y2 ~~ y2 + y2 + y3
    y3 ~~ y3 + y4
    y4 ~~ y4

    # all variances and covariances are freely estimated
'
```

Fit the model:

```{r}
model1fit <- lavaan::sem(model1, data = Data, 
                         cluster = "clus",
                         estimator = 'MLR',
                         fixed.x = FALSE)

lavaan::summary(model1fit, fit.measures = T, standardized = T)
```

<!------------------------------>
## PART II: Multilevel CFA 2: Between-only construct
<!------------------------------>

+ Example: construct reflects self-reported ‘school climate’ measured by a questionnaire filled in by the school principles
+ We will only have one response for each school 
+ We may collect other variables from students/teachers in the schools though

Note that the following model syntax:

```{r, eval=FALSE}
model2.wrong <- '
  level: 1
  # perhaps other level-1 variables

  level: 2
    fb =~ y1 + y2 + y3 + y4
'

model2fit <- lavaan::sem(model2.wrong, data = Data, 
                         cluster = "clus",
                         estimator = 'MLR',
                         fixed.x = FALSE)
```

```{r, eval=FALSE}
Error in lav_partable_vnames(tmp.lav, type = "ov", level = tmp.level.values[l]) : lavaan ERROR: level column does not contain value `1'
```

won't work because there is nothing at level 1. Instead, specify this model just like a regular CFA model:

```{r}
model2 <- '
    fb =~ y1 + y2 + y3 + y4
'
```


Fit the model:

```{r}
model2fit <- lavaan::sem(model2, data = Data, 
                         cluster = "clus",
                         estimator = 'MLR',
                         fixed.x = FALSE)

lavaan::summary(model2fit, fit.measures = T, standardized = T)
```


<!------------------------------>
## PART III: Multilevel CFA 3: Shared cross-level construct
<!------------------------------>

+ This model specifies the latent variable both at the within and the between level;
+ However, the latent variable only makes sense at the between level so SEM model is only built at the between level;
+ The indicators are correlated at the within level; 

```{r m3}
model3 <- '
  level: 1
    y1 ~~ y1 + y2 + y3 + y4
    y2 ~~ y2 + y3 + y4
    y3 ~~ y3 + y4
    y4 ~~ y4

  level: 2
    fs =~ y1 + y2 + y3 + y4

 # Fix Significant Heywood Cases
  y2 ~~ v2*y2
  y4 ~~ v4*y4
  v2 > 0
  v4 > 0 
'
```

Fit the model:

```{r m3fit}
model3fit <- lavaan::sem(model3, data = Data, 
                         cluster = "clus",
                         estimator = 'MLR',
                         fixed.x = FALSE)

lavaan::summary(model3fit, standardized = T, fit.measures = T)
```


<!------------------------------>
## PART IV: Multilevel CFA 4: Configural construct
<!------------------------------>

+ Model 4a specifies the latent variable both at the within and the between level;
+ The CFA at each level should have the same factor structure, but not necessarily the same parameter estimates;

```{r}
model4a <- '
  level: 1
    fw =~ y1 + y2 + y3 + y4

  level: 2
    fb =~ y1 + y2 + y3 + y4
'

model4afit <- lavaan::sem(model4a, data = Data, 
                         cluster = "clus",
                         estimator = 'MLR',
                         fixed.x = FALSE)

lavaan::summary(model4afit, fit.measures = T, standardized = T)
```

+ Model 4b specifies the same CFA at each level and requires the same factor loadings;

```{r}
model4b <- '
  level: 1
    fw =~ a*y1 + b*y2 + c*y3 + d*y4

  level: 2
    fb =~ a*y1 + b*y2 + c*y3 + d*y4
'

model4bfit <- lavaan::sem(model4b, data = Data, 
                         cluster = "clus",
                         estimator = 'MLR',
                         fixed.x = FALSE)

lavaan::summary(model4bfit, fit.measures = T, standardized = T)
```


<!------------------------------>
## PART V: Multilevel CFA 5: Shared + Configural construct
<!------------------------------>

+ This model specifies the latent variable both at the within and the between level;
+ The CFA at each level should have the same factor structure, but not necessarily the same parameter estimates;

```{r}
model5 <- '
  level: 1
    fw =~ a*y1 + b*y2 + c*y3 + d*y4

  level: 2
    fb =~ a*y1 + b*y2 + c*y3 + d*y4   # configural
    fs =~ y1 + y2 + y3 + y4           # shared
    
    # fb and fs must be orthogonal
    fs ~~ 0*fb
'

model5fit <- lavaan::sem(model5, data = Data, 
                         cluster = "clus",
                         estimator = 'MLR',
                         fixed.x = FALSE)

lavaan::summary(model5fit, fit.measures = T, standardized = T)
```


<!------------------------------>
## PART VI: Model Comparison 
<!------------------------------>

```{r}
m1 = fitMeasures(model1fit, fit.measures = c("chisq.scaled", "df.scaled", "pvalue.scaled", "rmsea.robust", "rmsea.ci.lower.robust", "rmsea.ci.upper.robust", "cfi.robust", "tli.robust", "srmr_within", "srmr_between"))
m2 = fitMeasures(model2fit, fit.measures = c("chisq.scaled", "df.scaled", "pvalue.scaled", "rmsea.robust", "rmsea.ci.lower.robust", "rmsea.ci.upper.robust", "cfi.robust", "tli.robust", "srmr_within", "srmr_between"))
m3 = fitMeasures(model3fit, fit.measures = c("chisq.scaled", "df.scaled", "pvalue.scaled", "rmsea.robust", "rmsea.ci.lower.robust", "rmsea.ci.upper.robust", "cfi.robust", "tli.robust", "srmr_within", "srmr_between"))
m4a = fitMeasures(model4afit, fit.measures = c("chisq.scaled", "df.scaled", "pvalue.scaled", "rmsea.robust", "rmsea.ci.lower.robust", "rmsea.ci.upper.robust", "cfi.robust", "tli.robust", "srmr_within", "srmr_between"))
m4b = fitMeasures(model4bfit, fit.measures = c("chisq.scaled", "df.scaled", "pvalue.scaled", "rmsea.robust", "rmsea.ci.lower.robust", "rmsea.ci.upper.robust", "cfi.robust", "tli.robust", "srmr_within", "srmr_between"))
m5 = fitMeasures(model5fit, fit.measures = c("chisq.scaled", "df.scaled", "pvalue.scaled", "rmsea.robust", "rmsea.ci.lower.robust", "rmsea.ci.upper.robust", "cfi.robust", "tli.robust", "srmr_within", "srmr_between"))
```
```{r,message=FALSE}
#install.packages('qpcR')
library(qpcR)
round(qpcR:::cbind.na(m1, m2, m3, m4a, m4b, m5), 3)
```

The final model goes to (drumroll)...model4b!

<!------------------------------>
## PART VII: Adding Covariates to Multilevel SEM
<!------------------------------>

### Model A: Adding a within-only covariate

```{r}
model4wCovA <- '
  level: 1
    fw =~ a*y1 + b*y2 + c*y3 + d*y4
    fw ~ x1

  level: 2
    fb =~ a*y1 + b*y2 + c*y3 + d*y4
'

model4wCovAfit <- lavaan::sem(model4wCovA, data = Data, 
                              cluster = "clus", 
                              estimator = 'MLR',
                              fixed.x = FALSE)

lavaan::summary(model4wCovAfit, fit.measures = T, standardized = T)
```

### Model B: Adding a between-only covariate

```{r}
model4wCovB <- '
  level: 1
    fw =~ a*y1 + b*y2 + c*y3 + d*y4

  level: 2
    fb =~ a*y1 + b*y2 + c*y3 + d*y4
    fb ~ w
'

model4wCovBfit <- lavaan::sem(model4wCovB, data = Data, 
                              cluster = "clus", 
                              estimator = 'MLR',
                              fixed.x = FALSE)

lavaan::summary(model4wCovBfit, fit.measures = T, standardized = T)
```

### Model C: Adding a covariate at both levels

```{r}
model4wCovC <- '
  level: 1
    fw =~ a*y1 + b*y2 + c*y3 + d*y4
    fw ~ x1

  level: 2
    fb =~ a*y1 + b*y2 + c*y3 + d*y4
    fb ~ x1
'

model4wCovCfit <- lavaan::sem(model4wCovC, data = Data, 
                              cluster = "clus", 
                              estimator = 'MLR',
                              fixed.x = FALSE)

lavaan::summary(model4wCovCfit, fit.measures = T, standardized = T)
```


<!------------------------------>
## PART VII: Final Model
<!------------------------------>

```{r}
modelFinal <- '
  level: 1
    fw =~ a*y1 + b*y2 + c*y3 + d*y4
    fw ~ x1 + x2

  level: 2
    fb =~ a*y1 + b*y2 + c*y3 + d*y4
    fb ~ w
'

modelFinalfit <- lavaan::sem(modelFinal, data = Data, 
                              cluster = "clus", 
                              estimator = 'MLR',
                              fixed.x = FALSE)

lavaan::summary(modelFinalfit, fit.measures = T, standardized = T)
```

